---
title: "s2_260120_qc_dimred"
author: "nlc and ZOD"
date: "20/01/2026"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```

<style>
    body .main-container {
        max-width: 2160px;
    }
</style>

# Experiment & contact info

PIs: Rosalie Sears (searsr@ohsu.edu), Laura Heiser (heiserl@ohsu.edu)

Sample preparation: Zinab Doha (dohaz@ohsu.edu) and (ztoha@taibahu.edu.sa)

Library prep from single cells: Xi Li & Colin Daniel (danielc@ohsu.edu)

Analysis: Nick Calistri (calistri@ohsu.edu)

Sequencing performed by OHSU MPSSR

## Analysis design

- load so_list.rds

- combine all libraries into so_merge

- process so_merge like normal (dimred & clustering)

- Compute qc metrics and calculate cellcycle score

- Analyze doublet calls vs cluster id to see if any clusters are majority doublet

- Filter based on QC metrics and sample ID

- Perform: Normalization, variablefeature selection, linear (PCA) and non linear (UMAP) dimensionality reduction using both standard and SCTransform workflows

- save filtered seurat objects as .rds files

# Set up

## Load environment from local project renv

```{r}
if(!dir.exists('renv')){
  print('Error - renv library not found')
}else{
  print('Loading packages from local renv library')
  library(renv)
  renv::load()
}
```

## Load libraries
```{r}
#library(Matrix)
library(tidyverse)
library(Seurat)
#library(rliger)
library(SeuratWrappers)
#library(ggalluvial)
library(harmony)
```

## load s1_so_list.rds

```{r}
#so_list <- readRDS('analysis_files/s1_seurat_list.rds')

so_files <- list.files('analysis_files/', pattern = '250505.rds') %>%
  grep(., pattern = '^s1', value = TRUE)

so_list <- list()
for(i in so_files){
  curr_so <- readRDS(paste0('analysis_files/', i))
  curr_lib_name <- unique(curr_so@meta.data$library_id)
  
  so_list[[curr_lib_name]] <- curr_so
  rm(curr_so)
}

gc()
```

## Mouse/tumor stats from README metadata file

```{r}
# Read in readme with HTO and tumor ID
lib_meta <- read_csv('original_data/221111_README.csv')

# Add the run_library variable
lib_meta <- lib_meta %>%
  mutate(run_library = paste0(mpssr_seq_run_name,
                              '_',
                              mpssr_library_name,
                              str_replace_na(library_suffix, replacement = '')))

lib_meta$mouse_id <- str_split(lib_meta$condition_string, pattern = '_', simplify = TRUE)[,1]

tumor_histology_dict <- lib_meta %>%
  dplyr::select(condition_string, histology) %>%
  distinct()
```

# Perform QC filtering and subsample to 500 cells per tumor per library

```{r}
for(i in 1:length(so_list)){
  curr_so <- so_list[[i]]
  
  # Add metadata: percent.mt, phenotype (from tumor ID)
  curr_so[['percent.mt']] <- PercentageFeatureSet(curr_so, pattern = '^mt-')
  
  curr_so[['percent.ribo']] <- PercentageFeatureSet(curr_so, pattern = '^Rp[sl][[:digit:]]')
  
  curr_so[['phenotype']] <- str_replace(str_split(curr_so@meta.data$tumor,
                                                   pattern = '_',
                                                   simplify = TRUE)[,1],
                                         '[:digit:]+',
                                         '')
  
  curr_so[['mouse']] <- str_split(curr_so@meta.data$tumor,
                      pattern = '_',
                      simplify = TRUE)[,1]
  
  curr_so[['histology']] <- plyr::mapvalues(x = curr_so@meta.data$tumor,
                                             from = tumor_histology_dict$condition_string,
                                             to = tumor_histology_dict$histology)

  # Filter to only cells with >250 features, <25% mitochondrial reads and assigned Singlet from DF and HTO demultiplex
  meta_filtered <- curr_so@meta.data %>%
    filter(nFeature_RNA > 300) %>%
    filter(nCount_RNA > 1000) %>%
    filter(percent.mt < 15) %>%
    filter(percent.ribo < 50) %>%
    filter(DF.classifications == 'Singlet') %>%
    filter(HTO_classification.global == 'Singlet')
  
  # rm(cells_to_keep)
  # for(j in 1:length(unique(meta_filtered$mouse))){
  #   curr_mouse <- meta_filtered %>%
  #     filter(mouse == unique(meta_filtered$mouse)[j])
  #   
  #   set.seed(i*j)
  #   
  #   curr_mouse_sub <- curr_mouse %>%
  #     slice_sample(n = min(500, nrow(curr_mouse)), replace = FALSE)
  #   
  #   
  #   if(j == 1){
  #     cells_to_keep <- row.names(curr_mouse_sub)
  #   }else{
  #     cells_to_keep <- c(cells_to_keep,
  #                        row.names(curr_mouse_sub))
  #   }
  #   
  # }
# 
#   curr_so <- subset(curr_so,
#                     cells = cells_to_keep)
  
  curr_so <- subset(curr_so,
                    cells = row.names(meta_filtered))

  so_list[[i]] <- curr_so
}
```


# Merge data and visualize QC metrics

## Merge seurat objects (no integration, simple appending)

```{r}
so_merge <- merge(x = so_list[[1]], y = unlist(so_list)[-1])
```

### Remove old seurat objects/lists and run GC

```{r}
so_list <- NULL
gc()
```

# QC stats table for supplement
```{r}
# Stats table
so_tumor_stats <- so_merge@meta.data %>%
  group_by(tumor) %>%
  summarize(nCells = n(),
            mean_nCount_RNA = mean(nCount_RNA),
            mean_nFeature_RNA = mean(nFeature_RNA),
            mean_percent.mt = mean(percent.mt)) 

so_tumor_stats %>%
  DT::datatable()

# Statistics used in results text

## Number of tumors
length(unique(so_merge@meta.data$tumor))

## Number of mice
length(unique(so_merge@meta.data$mouse))

## Number of unique libraries
length(unique(so_merge@meta.data$library_id))

## Mean and range number of cells per tumor
range(so_tumor_stats$nCells)
sum(so_tumor_stats$nCells)

## Mean genes per cell
mean(so_merge@meta.data$nFeature_RNA)
mean(so_merge@meta.data$nCount_RNA)

## Number of cells per tumor
so_merge@meta.data %>%
  dplyr::select(tumor, phenotype) %>%
  table()

```

# Filter to samples of interest, run standard Seurat dimred and save

## Filter samples of interest

```{r}
# Subset to samples of interest
#so_filt <- subset(so_merge, subset = phenotype #%in% c('V', 'R', 'S'))

# Remove unneeded so_merge object and clean up
#rm(so_merge)
#gc()
```

## Standard seurat analysis (log normalization)

```{r}
so_filt <- NormalizeData(so_merge)
so_filt <- FindVariableFeatures(so_filt,
                                nfeatures = 5000)
so_filt <- ScaleData(so_filt)
so_filt <- RunPCA(so_filt)
ElbowPlot(so_filt, ndims = 50)

so_filt <- RunUMAP(so_filt, dims = 1:20)
```



## Save results

```{r}
saveRDS(so_filt, file = 'analysis_files/s2_vsr_standard.rds')
```


## Visualize output

```{r, fig.asp = 1}
DimPlot(so_filt,
        group.by = 'library_id')+
  coord_equal()

DimPlot(so_filt,
        group.by = 'tumor')+
  coord_equal()

DimPlot(so_filt,
        group.by = 'phenotype', split.by = 'phenotype')+
  coord_equal()

FeaturePlot(so_filt,
            features = c('nFeature_RNA', 'nCount_RNA'),
            coord.fixed = TRUE)
```

```{r, fig.asp = 3}
DimPlot(so_filt,
        label = FALSE,
        group.by = 'tumor')+
  coord_equal()+
  facet_wrap(~tumor, ncol = 3)+
  theme(legend.position = 'none')
```

## Save results

```{r}
saveRDS(so_filt, file = 'analysis_files/s2_vsr_standard.rds')
```

# Run SCT

## Run SCT normalization

```{r}
so_sct <- SCTransform(so_filt,
                      method = 'glmGamPoi',
                      variable.features.n = 5000,
                      vars.to.regress = c('percent.mt', 'percent.ribo'))

# Remove so_filt
rm(so_filt)
gc()
```

## Dimred on SCT normalized

```{r}
so_sct <- ScaleData(so_sct)
so_sct <- RunPCA(so_sct)
ElbowPlot(so_sct, ndims = 50)

so_sct <- RunUMAP(so_sct, dims = 1:20)
```


## Visualize output

```{r, fig.asp = 1}
DimPlot(so_sct,
        group.by = 'library_id')+
  coord_equal()

DimPlot(so_sct,
        group.by = 'tumor')+
  coord_equal()

DimPlot(so_sct,
        group.by = 'phenotype', split.by = 'phenotype')

FeaturePlot(so_sct,
            features = c('nFeature_RNA', 'nCount_RNA'),
            coord.fixed = TRUE)
```


```{r, fig.asp = 3}
DimPlot(so_sct,
        label = FALSE,
        group.by = 'tumor')+
  coord_equal()+
  facet_wrap(~tumor, ncol = 3)+
  theme(legend.position = 'none')
```

## Save results

```{r}
saveRDS(so_sct, file = 'analysis_files/s2_vsr_sct_250505.rds')
```

# Harmony integration

```{r}
# so_sct <- readRDS(file = 'analysis_files/s2_vsr_sct_250505.rds')
# 
# so_sct@meta.data$batch_id <- str_split(so_sct@meta.data$library_id,
#                                     pattern = '_',
#                                     simplify = TRUE)[,1]
# 
# # so_sct_split <- so_sct[["SCT"]] <- split(so_sct[["SCT"]], f = so_sct$batch)
# # 
# # so_harmony <- IntegrateLayers(object = so_sct_split,
# #                               method = HarmonyIntegration,
# #                               orig.reduction = "pca",
# #                               new.reduction = 'harmony',
# #                               assay = "SCT",
# #                               verbose = TRUE)
# 
# so_harmony <- harmony::RunHarmony(so_sct,
#                          group.by.vars = 'batch_id',
#                          reduction = 'pca',
#                          assay.use = 'SCT',
#                          project.dim = FALSE,
#                          verbose = TRUE)
# 
# saveRDS(so_harmony, file = 'analysis_files/s2_vsr_harmony_250505.rds')

```


# sessionInfo()

```{r}
sessionInfo()
```

