---
title: "s3_260120_cluster_optimization"
author: "nlc and ZOD"
date: "20/01/2026"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```

<style>
    body .main-container {
        max-width: 2160px;
    }
</style>

# Experiment & contact info

PIs: Rosalie Sears (searsr@ohsu.edu), Laura Heiser (heiserl@ohsu.edu)

Sample preparation: Zinab Doha (dohaz@ohsu.edu) AND (ztoha@taibahu.edu.sa)

Library prep from single cells: Xi Li & Colin Daniel (danielc@ohsu.edu)

Analysis: Nick Calistri (calistri@ohsu.edu)

Sequencing performed by OHSU MPSSR

## Analysis design

- load merged SO

- Optimize leiden clustering for maximized silhouette width

- Visualize cluster versus lineage markers

- Manually assign lineage

- Run 3D UMAP and visualize clusters + lineage in 3d

- Find DEGs (upregulated) for each cluster compared to all other clusters

- Save updated so_merge in updated .rds file


# Set up

## Load environment from local project renv

```{r}
if(!dir.exists('renv')){
  print('Error - renv library not found')
}else{
  print('Loading packages from local renv library')
  library(renv)
  renv::load()
}
```


## Load libraries
```{r}
library(tidyverse)
library(Seurat)
library(cluster)
library(bluster)
# library(future)
# plan("multisession",
#      workers = 6)
```

## load s2_vehicle_integrated.rds and set seed
```{r}
so_merge <- readRDS('analysis_files/s2_vsr_standard.rds')

so_merge@meta.data$cell_barcode <- rownames(so_merge@meta.data)

set.seed(3)
```

## Write cluster optimization functionfunction

```{r}
cluster_qc <- function(seurat_object, embedding, ndims = ncol(embedding)){
  
  unique_clusterings <- grep(colnames(seurat_object@meta.data),
                                 pattern = '*_res.',
                                 value = TRUE)
  
  clust_assignments <- seurat_object@meta.data %>%
    dplyr::select(all_of(unique_clusterings))
  
  
  for(i in 1:ncol(clust_assignments)){
    
    curr_resolution <- str_remove(unique_clusterings[i],
                                  pattern = paste0(DefaultAssay(seurat_object), '_snn_res.')) %>%
      as.numeric()
    
    p1 <- DimPlot(seurat_object,
                  group.by = unique_clusterings[i],
                  label = TRUE)+
      coord_equal()+
      ggtitle(unique_clusterings[i])
    
    print(p1)
    
    curr_clusters <- clust_assignments[,i]
    
    sil <- approxSilhouette(x = embedding,
                            clusters = curr_clusters)
    
    boxplot(split(sil$width, curr_clusters),
            main = paste0('Resolution: ', curr_resolution, '\n Mean sil.width: ', mean(sil$width)))
    
    best.choice <- ifelse(sil$width > 0,
                          curr_clusters,
                          sil$other)
    
    table(Assigned=curr_clusters, Closest=best.choice)
    
    # 
    rmsd <- clusterRMSD(embedding, curr_clusters)
    barplot(rmsd,
            main = paste0('Resolution: ', curr_resolution, '\n Mean rmsd: ', mean(rmsd)))
    
    
    curr_qc <- tibble(res = curr_resolution,
                   n_clusters = length(unique(sil$cluster)),
                   mean_sil_width = mean(sil$width),
                   mean_rmsd = mean(rmsd))
    
    if(i == 1){
      qc <- curr_qc
    }else{
      qc <- rbind(qc, curr_qc)
    }
    
  }
  return(qc)
}
```

```{r}
ElbowPlot(so_merge, ndims = 50)
reduction_to_use <- 'pca'
ndims <- 15

###

so_merge <- FindNeighbors(so_merge,
                          reduction = reduction_to_use,
                          dims = 1:ndims)

so_merge <- RunUMAP(so_merge,
                    reduction = reduction_to_use,
                    dims = 1:ndims)


DimPlot(so_merge)+
  coord_equal()
```

## Sweep cluster resolution

```{r}
so_merge <- FindClusters(
  object = so_merge,
  resolution = seq(from = 0.05, to = 0.75, by = 0.05),
  algorithm = 1
)

```

## Compute {sil_width, RMSD} and visualize

```{r}
# Visualize results for each resolution
so_cluster_qc <- cluster_qc(seurat_object = so_merge,
                            embedding = so_merge@reductions$pca@cell.embeddings[,1:ndims])

# Select highest silhouette width
highest_res <- so_cluster_qc %>%
  filter(mean_sil_width/mean_rmsd == max(mean_sil_width/mean_rmsd)) %>%
  dplyr::select(res) %>%
  pull()

highest_res <- 0.25
  
# Gather for plotting
cluster_qc_gathered <- so_cluster_qc %>%
  gather(- c(res), value = 'value', key = 'metric')

# Visualize aggregated results
ggplot(cluster_qc_gathered, aes(x = res, y = value, color = metric))+
  geom_point()+
  geom_line()+
  theme_bw()+
  facet_wrap(~metric, ncol = 3, scales = 'free_y')+
  scale_x_continuous(breaks = seq(from = 0, to = 2.5, by = 0.1))+
  xlab('Louvain resolution')+
  geom_vline(xintercept = highest_res, linetype = 'dashed')+
  RotatedAxis()

ggplot(so_cluster_qc, aes(x = mean_sil_width, y = mean_rmsd, label = res, color = n_clusters))+
  geom_point()+
  ggrepel::geom_text_repel()+
  theme_bw()+
  RotatedAxis()

ggplot(so_cluster_qc, aes(x = n_clusters, y = mean_sil_width/mean_rmsd, label = res, color = n_clusters))+
  geom_point()+
  ggrepel::geom_text_repel()+
  theme_bw()+
  RotatedAxis()
```

## Save optimal clustering value as 'cluster_l1' and remove the remainder

```{r}
so_merge <- FindClusters(so_merge,
                         algorithm = 1,
                         resolution = 0.25)
```
```{r}
DimPlot(so_merge, label = TRUE, label.size = 5)+
  coord_equal()+
  theme(legend.position = 'none')
```

```{r}
ggplot(so_merge@meta.data, aes(x = seurat_clusters, fill = tumor))+
  geom_bar()+
  theme_bw()+
  facet_wrap(~tumor)
```


```{r}
# Store optimal clustering
so_merge@meta.data$cluster_l1 <- so_merge@meta.data[,paste0('RNA_snn_res.', highest_res)]

so_merge@meta.data$seurat_clusters <- so_merge@meta.data$cluster_l1

Idents(so_merge) <- 'cluster_l1'

DimPlot(so_merge,
        group.by = 'cluster_l1',
        label = TRUE)+
  coord_equal()+
  theme(legend.position = 'none')

# Remove the remainder
so_merge@meta.data <- so_merge@meta.data %>%
  dplyr::select(-grep(colnames(so_merge@meta.data),
                      pattern = '^SCT_snn_res*',
                      value = TRUE))

```

## Create csv dictionary for cluster number <-> lineage (if it doesn't already exist)

```{r}
if(file.exists('s3_cluster_lineage.csv')){
  print('Cluster lineage dictionary already exists')
}else{
  print('Created blank cluster-lineage dictionary (s3_cluster_lineage.csv)')
  clust_lin <- tibble(cluster = sort(unique(so_merge@meta.data$cluster_l1)),
                      lineage = '')
  write_csv(clust_lin, file = 's3_cluster_lineage.csv')
}
```

# Visualize lineage markers by cluster

```{r}
lin_markers <- c('Ptprc', 'Cd14', 'Cd74', 'Itgax', 'S100a8', 'Cd3e', 'Cd4', 'Cd8a', 'Foxp3', 'Trac', 'Trdc', 'Cd79a', 'Igkc', 'Epcam', 'Cdh1', 'Krt5', 'Pdgfra', 'Pdgfrb', 'Acta2', 'Col1a1', 'Pecam1', 'Mki67')

DotPlot(so_merge,
        features = lin_markers,
        cols = 'RdYlBu')+
  theme_bw()+
  coord_flip()
```

# Assign lineage from csv

```{r}
lineage_dict <- read_csv('s3_cluster_lineage.csv')

so_merge[['lineage']] <- plyr::mapvalues(x = so_merge@meta.data$ cluster_l1,
                                         from = lineage_dict$cluster,
                                         to = lineage_dict$lineage)

p1 <- DimPlot(so_merge,
        group.by = 'cluster_l1',
        label = TRUE,
        label.size = 5)+
  coord_equal()+
  theme(legend.position = 'none')

p2 <- DimPlot(so_merge,
        group.by = 'lineage')+
  coord_equal()


gridExtra::grid.arrange(p1, p2, ncol = 2)
```
```{r}
DimPlot(so_merge,
        group.by = 'phenotype')+
  coord_equal()
```


# Look at lineage markers vs cluster/lineage

```{r, fig.asp = 0.8}

DotPlot(so_merge,
        features = lin_markers,
        cols = 'RdYlBu',
        group.by = 'cluster_l1')+
  theme_bw()+
  coord_flip()

DotPlot(so_merge,
        features = lin_markers,
        cols = 'RdYlBu',
        group.by = 'lineage')+
  coord_flip()+
  RotatedAxis()
```

# Cluster relationship investigation

## Lineage vs mean embedding

```{r, fig.asp = 1}
# Average iNMF embeddings to show relationship to lineage
cluster_embedding <- tibble(lineage = so_merge@meta.data$lineage) %>%
  cbind(so_merge@reductions$pca@cell.embeddings) %>%
  group_by(lineage) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(data.frame(cluster_embedding[,2:ncol(cluster_embedding)],
                              row.names = cluster_embedding$lineage),
                   main = 'Average pca embedding per cluster \n column z-scored',
                   breaks = seq(from = -2.5, to = 2.5, length.out = 100))

# Average pca embedding to show relationship to cluster, annotated by lineage
cluster_embedding <- tibble(cluster_l1 = so_merge@meta.data$cluster_l1) %>%
  cbind(so_merge@reductions$pca@cell.embeddings) %>%
  group_by(cluster_l1) %>%
  summarize(across(everything(), list(mean)))

annotation_rows <- data.frame(lineage = lineage_dict$lineage,
                              row.names = lineage_dict$cluster)

pheatmap::pheatmap(data.frame(cluster_embedding[,2:ncol(cluster_embedding)],
                              row.names = cluster_embedding$cluster_l1),
                   main = 'Average pca embedding per cluster',
                   breaks = seq(from = -2.5, to = 2.5, length.out = 100),
                   annotation_row = annotation_rows)
```

# Correlation between clusters

code adapted from: https://github.com/satijalab/seurat/issues/1552

```{r, fig.asp = 1}
Idents(so_merge) <- 'cluster_l1'

av.exp <- AverageExpression(so_merge)$RNA[VariableFeatures(so_merge),]
cor.exp <- as.data.frame(cor(as.matrix(av.exp), method = 'pearson'))

cor_anno <- annotation_rows
row.names(cor_anno) <- paste0('g', row.names(cor_anno))

pheatmap::pheatmap(cor.exp,
                   annotation_row = cor_anno,
                   annotation_col = cor_anno)

DimPlot(so_merge,
        label = TRUE)+
  coord_equal()
```


# Celltype fractions

## Visualize celltype fractions per tumor
```{r, fig.asp = 1}

so_merge_meta <- so_merge@meta.data %>%
  mutate(tumor_pheno = paste0(tumor, '-', phenotype))

freq_table <- so_merge_meta %>%
  dplyr::select(lineage, tumor_pheno) %>%
  table()

col_anno <- data.frame(log10_n_cells = log10(colSums(freq_table)),
                       row.names = colnames(freq_table),
                       tumor_pheno = colnames(freq_table))

tumor_pheno_hist <- so_merge_meta %>%
  dplyr::select(tumor_pheno, histology, phenotype) %>%
  distinct()

col_anno <- left_join(x = col_anno,
                      y = tumor_pheno_hist,
                      by = 'tumor_pheno',
                      suffix = c('', ''))

col_anno <- data.frame(log10_n_cells = col_anno$log10_n_cells,
                       histology = col_anno$histology,
                       phenotype = col_anno$phenotype,
                       row.names = col_anno$tumor_pheno)



pheatmap::pheatmap(prop.table(freq_table, margin = 2), 
                   display_numbers = freq_table,
                   main = 'celltype fraction \n Number = n_cells per celltype per tumor',
                   annotation_col = col_anno)

```

```{r, fig.asp = 1.5}
# By cluster

freq_table <- so_merge_meta %>%
  dplyr::select(cluster_l1, tumor_pheno) %>%
  table()

row_anno <- data.frame(row.names = lineage_dict$cluster,
                       lineage = lineage_dict$lineage)

pheatmap::pheatmap(prop.table(freq_table, margin = 2), 
                   display_numbers = freq_table,
                   main = 'celltype fraction \n Number = n_cells per celltype per tumor',
                   annotation_col = col_anno,
                   annotation_row = row_anno)

# PCA of composition
props <- prop.table(freq_table, margin = 2) %>%
  data.frame() %>%
  spread(key = tumor_pheno,
         value = Freq) %>%
  data.frame(.,
             row.names = 'cluster_l1')

props.pca <- prcomp(t(props))

tumor_meta <- so_merge@meta.data %>%
  dplyr::select(tumor, library_id, histology) %>%
  distinct()

pca_tib <- as_tibble(props.pca$x) %>%
  mutate(tumor.pheno = row.names(props.pca$x)) %>%
  mutate(tumor = str_split(tumor.pheno,
                           pattern = '\\.',
                           simplify = TRUE)[,1],
         phenotype = str_split(tumor.pheno,
                               pattern = '\\.',
                               simplify = TRUE)[,2]) %>%
  left_join(x = .,
            y = tumor_meta,
            by = 'tumor')

summary(props.pca)

ggplot(pca_tib, aes(x = PC1, y = PC2, color = phenotype, label = tumor))+
  geom_point()+
  geom_text()+
  coord_equal()+
  theme_bw()

ggplot(pca_tib, aes(x = PC1, y = PC2, color = library_id, label = tumor))+
  geom_point()+
  geom_text()+
  coord_equal()+
  theme_bw()

ggplot(pca_tib, aes(x = PC1, y = PC2, color = histology, label = tumor))+
  geom_point()+
  geom_text()+
  coord_equal()+
  theme_bw()

```

# 3d umap

```{r, fig.asp = 1}
so_merge <- RunUMAP(so_merge,
                    reduction = 'pca',
                    dims = 1:10,
                    n.components = 3L,
                    reduction.name = 'umap3d',
                    reduction.key = 'umap3d_')

so_meta <- so_merge@meta.data %>%
  mutate(umap3d_1 = so_merge@reductions$umap3d@cell.embeddings[,1]) %>%
  mutate(umap3d_2 = so_merge@reductions$umap3d@cell.embeddings[,2]) %>%
  mutate(umap3d_3 = so_merge@reductions$umap3d@cell.embeddings[,3]) %>%
  mutate(cluster_l1 = as.factor(cluster_l1))

plotly::plot_ly(so_meta,
                x = ~umap3d_1,
                y = ~umap3d_2,
                z = ~umap3d_3,
                color = ~cluster_l1,
                alpha = 0.9)

plotly::plot_ly(so_meta,
                x = ~umap3d_1,
                y = ~umap3d_2,
                z = ~umap3d_3,
                color = ~lineage)
```

# Find DEG for each cluster at global (across all clusters) level

## Find DEG

```{r, fig.asp = 1.5}
Idents(so_merge) <- 'cluster_l1'

#so_merge <- JoinLayers(so_merge)

#so_merge <- PrepSCTFindMarkers(so_merge)

markers <- FindAllMarkers(so_merge)

top_markers <- markers %>%
  group_by("cluster_l1") %>%
  filter(p_val_adj < 0.05 & pct.1 > 0.2) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 3)

# Scale data by all DEGs
so_merge <- ScaleData(so_merge, 
                      features = unique(top_markers$gene))

DoHeatmap(so_merge, features = unique(top_markers$gene))

DotPlot(so_merge,
        features = unique(top_markers$gene),
        cols = 'RdYlBu')+
  coord_flip()+
  theme_bw()+
  RotatedAxis()

write_csv(markers, 'analysis_files/s3_cluster_l1_markers.csv')
```

# save rds output

```{r}
saveRDS(so_merge, file = 'analysis_files/s3_lineages_250505.rds')
```

# sessionInfo()

```{r}
sessionInfo()
```

