---
title: "s4_260120_celltype_definition"
author: "nlc and zod"
date: "20/01/2026"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 1)
```

<style>
    body .main-container {
        max-width: 2160px;
    }
</style>

# Experiment & contact info

PIs: Rosalie Sears (searsr@ohsu.edu), Laura Heiser (heiserl@ohsu.edu)

Sample preparation: Zinab Doha (dohaz@ohsu.edu) and (ztoha@taibahu.edu.sa)

Library prep from single cells: Xi Li & Colin Daniel (danielc@ohsu.edu)

Analysis: Nick Calistri (calistri@ohsu.edu)

Sequencing performed by OHSU MPSSR

## Analysis design

- load seurat object

- separate each lineage with multiple clusters (epithelial, stromal, lymphoid, myeloid)

- reprocess within the context of the lineage (varfeatures, scale, dimred, cluster)

- perform DE within each lineage

- Convert cluster to celltype and annotate with relevant pathway/expression 

- Save annotated seurat obejct (s4_so_merge.rds) as well as the lineage subset seurat objects


### Object names

epi = epithelial
sni = stromal non-immune
lym = lymphoid
mye = myeloid

# Set up

## Load environment from local project renv

```{r}
if(!dir.exists('renv')){
  print('Error - renv library not found')
}else{
  print('Loading packages from local renv library')
  library(renv)
  renv::load()
  #reticulate::use_condaenv(condaenv = 'seurat')
}
```



## Load libraries
```{r}
library(tidyverse)
library(Seurat)
library(harmony)
library(bluster)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichR)
# library(UpSetR)
# library(eulerr)
```

## load s3_celltypes.rds and create batch variable

```{r}
so_merge <- readRDS('analysis_files/s3_lineages_250505.rds')

# Consider batch the vial of tubes that was split across two 10X lanes (prefix of library_id)

so_merge@meta.data$batch <- str_split(so_merge@meta.data$library_id,
                                      pattern = '_',
                                      simplify = TRUE)[,1]

# Remove any previous clustering attempts
so_merge@meta.data <- so_merge@meta.data %>%
  dplyr::select(-grep(colnames(so_merge@meta.data),
                      pattern = '*snn_res*',
                      value = TRUE))

```

## Create subset seurat objects

```{r}
so_epi <- subset(so_merge, subset = lineage == 'epithelial')

so_lym <- subset(so_merge, subset = lineage == 'lymphoid')

so_mye <- subset(so_merge, subset = lineage == 'myeloid')

so_str <- subset(so_merge, subset = lineage == 'stromal')

# Set default ndims
ndims <- 10
```

## Prepare subset processing function

```{r}
process_subset <- function(seurat_object,
                        ndims = 10,
                        nVarFeats = 3000){
  
  # DefaultAssay(seurat_object) <- 'RNA'
  # 
  # seurat_object <- NormalizeData(seurat_object)
  # seurat_object <- FindVariableFeatures(seurat_object,
  #                                       nfeatures = nVarFeats)
  # seurat_object <- ScaleData(seurat_object,
  #                            vars.to.regress = c('percent.mt', 'percent.ribo'))
  # 
  # seurat_object <- RunPCA(seurat_object)
  # 
  # # UMAP
  # seurat_object <- RunUMAP(seurat_object,
  #                     dims = 1:ndims,
  #                     n.neighbors = 50)
  # 
  # seurat_object <- FindNeighbors(seurat_object,
  #                                dims = 1:ndims)
  # 
  # seurat_object <- CellCycleScoring(seurat_object,
  #                                 s.features = cc.genes.updated.2019$s.genes,
  #                                 g2m.features = cc.genes.updated.2019$g2m.genes)
  
  # new 05/06/25
  seurat_object <- SCTransform(seurat_object,
                               variable.features.n = nVarFeats,
                               vars.to.regress = c('percent.mt', 'percent.ribo'))
  
  seurat_object <- RunPCA(seurat_object)
  
  seurat_object <- RunHarmony(seurat_object, group.by.vars = 'batch')
  
  seurat_object <- RunUMAP(seurat_object,
                      dims = 1:ndims,
                      n.neighbors = 100,
                      min_dist = 0.5,
                      reduction = 'harmony')
  
  seurat_object <- FindNeighbors(seurat_object,
                                 dims = 1:ndims,
                                 reduction = 'harmony')
  
  seurat_object <- CellCycleScoring(seurat_object,
                                  s.features = cc.genes.updated.2019$s.genes,
                                  g2m.features = cc.genes.updated.2019$g2m.genes)
  
  return(seurat_object)
}
```

## Prepare harmony subset processing function

```{r}
process_harmony_subset <- function(seurat_object,
                        ndims = 10,
                        nVarFeats = 3000){
  
  # Preprocess each batch separately
  # seurat_object[["RNA"]] <- split(seurat_object, f = seurat_object$batch)
  # seurat_object <- SCTransform(seurat_object,
  #                              vars.to.regress = c('percent.mt', 'percent.ribo'))
  # 
   DefaultAssay(seurat_object) <- 'SCT'

   seurat_object <- NormalizeData(seurat_object)
   seurat_object <- FindVariableFeatures(seurat_object)
   seurat_object <- ScaleData(seurat_object)
  
  seurat_object <- RunPCA(seurat_object)

  # Integrate batch with Harmony
  seurat_object <- IntegrateLayers(object = seurat_object,
                                   method = HarmonyIntegration,
                                   orig.reduction = "pca",
                                   new.reduction = 'harmony',
                                   verbose = FALSE)
  
  # Compute neighborhood and SNN from harmony dimred
  seurat_object <- RunUMAP(seurat_object,
                           reduction = 'harmony',
                           dims = 1:ndims)
  
  seurat_object <- FindNeighbors(seurat_object,
                                 reduction = 'harmony',
                                 dims = 1:ndims)
  
  # Join layers back together
  seurat_object <- JoinLayers(seurat_object)
  
  seurat_object <- PrepSCTFindMarkers(seurat_object)
  
  # Run cell cycle scoring  
  seurat_object <- CellCycleScoring(seurat_object,
                                    s.features = cc.genes.updated.2019$s.genes,
                                    g2m.features = cc.genes.updated.2019$g2m.genes)
  
  return(seurat_object)
}
```


## Prepare cluster optimization function

```{r}
cluster_qc <- function(seurat_object, embedding, ndims = ncol(embedding)){
  
  unique_clusterings <- grep(colnames(seurat_object@meta.data),
                                 pattern = '*_res.',
                                 value = TRUE)
  
  clust_assignments <- seurat_object@meta.data %>%
    dplyr::select(all_of(unique_clusterings))
  
  
  for(i in 1:ncol(clust_assignments)){
    
    curr_resolution <- str_remove(unique_clusterings[i],
                                  pattern = paste0(DefaultAssay(seurat_object), '_snn_res.')) %>%
      as.numeric()
    
    p1 <- DimPlot(seurat_object,
                  group.by = unique_clusterings[i],
                  label = TRUE)+
      coord_equal()+
      ggtitle(unique_clusterings[i])
    
    print(p1)
    
    curr_clusters <- clust_assignments[,i]
    
    sil <- approxSilhouette(x = embedding,
                            clusters = curr_clusters)
    
    boxplot(split(sil$width, curr_clusters),
            main = paste0('Resolution: ', curr_resolution, '\n Mean sil.width: ', mean(sil$width)))
    
    best.choice <- ifelse(sil$width > 0,
                          curr_clusters,
                          sil$other)
    
    table(Assigned=curr_clusters, Closest=best.choice)
    
    # 
    rmsd <- clusterRMSD(embedding, curr_clusters)
    barplot(rmsd,
            main = paste0('Resolution: ', curr_resolution, '\n Mean rmsd: ', mean(rmsd)))
    
    
    curr_qc <- tibble(res = curr_resolution,
                   n_clusters = length(unique(sil$cluster)),
                   mean_sil_width = mean(sil$width),
                   mean_rmsd = mean(rmsd))
    
    if(i == 1){
      qc <- curr_qc
    }else{
      qc <- rbind(qc, curr_qc)
    }
    
  }
  return(qc)
}
```

## Prepare DEG function

```{r}
subset_ego <- function(curr_sub, curr_markers){

  if(nrow(curr_markers) > 1){
    
    curr_top <- curr_markers %>%
      filter(avg_log2FC > 0.5) %>%
      filter(pct.1 > 0.1) %>%
      group_by(cluster) %>%
      arrange(desc(avg_log2FC)) %>%
      filter(p_val_adj <= 0.05) %>%
      slice_head(n = 5)
    
    DoHeatmap(curr_sub, features = curr_top$gene)
    
    for(i in sort(unique(curr_markers$cluster))){
      
      
      clust_markers <- curr_markers %>%
        filter(cluster == i)
      
      print(paste0('Evaluating cluster: ', i))
      
      volcano_plot <- ggplot(clust_markers, aes(x = avg_log2FC, y = -log10(p_val_adj), label = gene))+
        geom_point()+
        ggrepel::geom_text_repel()+
        ggtitle(paste0('Cluster: ', i))
      
      print(volcano_plot)
      
      curr_top <- curr_markers %>%
        filter(cluster == i) %>%
        filter(avg_log2FC >= 0.5) %>%
        filter(p_val_adj <= 0.05)
      
      curr_bottom <- curr_markers %>%
        filter(cluster == i) %>%
        filter(avg_log2FC <= -0.5) %>%
        filter(p_val_adj <= 0.05)
      
      # EnrichR celltype using Azimuth 2023 database
      er_celltype <- enrichR::enrichr(genes = curr_top$gene,
                                      databases = 'Azimuth_2023')$Azimuth_2023 %>%
        mutate(cluster = i)
      
      
      if(if(!is.null(er_celltype)){nrow(er_celltype) > 0}){
        plotEnrich(er_celltype, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value")
      
        print(head(er_celltype))
      }
      
      # Enrichment of gene ontology
      ego_top <- enrichGO(gene = curr_top$gene,
                OrgDb = org.Mm.eg.db,
                keyType = 'SYMBOL',
                ont = "ALL",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05)
  
      if(if(!is.null(ego_top)){nrow(ego_top) > 0 & nrow(curr_top) >= 3}){
        ego_top_genelist <- curr_top$avg_log2FC
        names(ego_top_genelist) <- curr_top$gene
        ego_top_genelist <- sort(ego_top_genelist, decreasing = TRUE)
        
        p1 <- dotplot(ego_top, title = paste0('Cluster ', i))
        
        p2 <- cnetplot(ego_top,
                       showCategory = 10)
        
        p3 <- heatplot(ego_top,
                       showCategory = 10)
        
        print(p1)
        print(p2)
        print(p3)
    }else(
      print('No upregulated GOs found')
    )
      
    
    }
      
  }else{
    print('FindAllMarkers did not return DEGs (only one cluster in celltype?)')
  }  
}
```

# Epithelial (04/17/26)

## Process epithelial data

##faster
```{r}
# Epithelial (05/25/25)

if (file.exists("analysis_files/s4_so_epithelial.rds")) {

  message("Epithelial subset already processed — loading")
  so_epi <- readRDS("analysis_files/s4_so_epithelial.rds")

} else {

  message("Processing epithelial subset (optimized)")

  DefaultAssay(so_epi) <- "RNA"

  # 1. Normalize (fast)
  so_epi <- NormalizeData(so_epi, verbose = FALSE)

  # 2. Variable features (LIMITED)
  so_epi <- FindVariableFeatures(
    so_epi,
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
  )

  # 3. Scale ONLY variable genes (huge speedup)
  so_epi <- ScaleData(
    so_epi,
    features = VariableFeatures(so_epi),
    verbose = FALSE
  )

  # 4. PCA (few dims first)
  so_epi <- RunPCA(
    so_epi,
    features = VariableFeatures(so_epi),
    npcs = 20,
    verbose = FALSE
  )

  # 5. Neighbors + clustering (lightweight)
  so_epi <- FindNeighbors(so_epi, dims = 1:10, verbose = FALSE)
  so_epi <- FindClusters(so_epi, resolution = 0.4, verbose = FALSE)

  # 6. UMAP (light)
  so_epi <- RunUMAP(so_epi, dims = 1:10, verbose = FALSE)

  saveRDS(so_epi, "analysis_files/s4_so_epithelial.rds")
}

```


```{r}

 if(file.exists('analysis_files/s4_so_epithelial.rds')){
   print('Epithelial subset already processed - loading saved file')
   so_epi <- readRDS('analysis_files/s4_so_epithelial.rds')
 }else{
   print('Preprocessed Epithelial seurat object not found, processing now. ETA <5 minutes')
   so_epi <- process_subset(so_epi,
                            ndims = 5)
 }

ElbowPlot(so_epi)

DimPlot(so_epi,
        group.by = 'phenotype')+
  coord_equal()+
  facet_wrap(~phenotype)

DimPlot(so_epi,
        group.by = 'cluster_l1')+
  coord_equal()

DimPlot(so_epi,
        group.by = 'Phase')+
  coord_equal()


```

## Optimize clusters

### Cluster sweep

```{r}
if(file.exists('analysis_files/s4_so_epithelial.rds')){
  print('Epithelial subset seurat file already exists, not running cluster optimization')
}else{
  
  # Sweep clusters
  so_epi <- FindClusters(
    object = so_epi,
    resolution = seq(from = 0.05, to = 0.7, by = 0.05),
    algorithm = 1
  )
  
  ndims <- 10
  
  so_cluster_qc <- cluster_qc(seurat_object = so_epi,
                              embedding = so_epi@reductions$harmony@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  best_res <- so_cluster_qc %>%
    filter(mean_sil_width/mean_rmsd == max(mean_sil_width/mean_rmsd)) %>%
    dplyr::select(res) %>%
    pull()
  
  p1 <- ggplot(cluster_qc_gathered, aes(x = res, y = value, color = metric))+
    geom_point()+
    geom_line()+
    theme_bw()+
    facet_wrap(~metric, ncol = 3, scales = 'free_y')+
    scale_x_continuous(breaks = seq(from = 0, to = 2.5, by = 0.5))+
    xlab('Louvain resolution')+
    geom_vline(xintercept = best_res, linetype = 'dashed')
  
  p2 <- ggplot(so_cluster_qc, aes(x = mean_sil_width, y = mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  p3 <- ggplot(so_cluster_qc, aes(x = n_clusters, y = mean_sil_width/mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  plot(p1)
  plot(p2)
  plot(p3)
  

}

best_res <- 0.25

```

### Remove swept clusters and set to optimal resolution

```{r, fig.asp = 1}
# if(file.exists('analysis_files/s4_so_epithelial.rds')){
#   print('Epithelial subset seurat file already exists, optimizing clustering resolution')
# }else{
#   
#   # Store optimal clustering
#   so_epi@meta.data$cluster_l2 <- paste0('epi_', so_epi@meta.data[,paste0('RNA_snn_res.', best_res)])
#   
#   so_epi@meta.data$seurat_clusters <- so_epi@meta.data$cluster_l2
#   
# 
#   
#   
#   
#   DimPlot(so_epi,
#           group.by = 'cluster_l2',
#           label = TRUE)+
#     coord_equal()+
#     theme(legend.position = 'none')
#   
#   # Remove the remainder
#   so_epi@meta.data <- so_epi@meta.data %>%
#     dplyr::select(-grep(colnames(so_epi@meta.data),
#                         pattern = '*snn_res*',
#                         value = TRUE))
# 
#   # Save processed object
#   saveRDS(so_epi, file = 'analysis_files/s4_so_epithelial.rds')
# }
# 
# # Visualize optimal sub
# 
# DimPlot(so_epi,
#         group.by = 'cluster_l2',
#         label = TRUE)+
#   coord_equal()
# 
# # Overlap
# 
# cluster_overlap <- so_epi@meta.data %>%
#   dplyr::select(cluster_l1, cluster_l2) %>%
#   table()
# 
# epi_nmi <- aricode::NMI(c1 = so_epi@meta.data$cluster_l1,
#                         c2 = so_epi@meta.data$cluster_l2)
```


```{r}
# Store optimal clustering (ALWAYS)
so_epi$cluster_l2 <- paste0(
  "epi_",
  so_epi@meta.data[[paste0("RNA_snn_res.", best_res)]]
)

so_epi$seurat_clusters <- so_epi$cluster_l2
Idents(so_epi) <- "cluster_l2"


DimPlot(
  so_epi,
  group.by = "cluster_l2",
  label = TRUE
) +
  coord_equal() +
  theme(legend.position = "none")


# Overlap

cluster_overlap <- so_epi@meta.data %>%
  dplyr::select(cluster_l1, cluster_l2) %>%
  table()

epi_nmi <- aricode::NMI(c1 = so_epi@meta.data$cluster_l1,
                        c2 = so_epi@meta.data$cluster_l2)

#saveRDS(so_epi, "analysis_files/s4_so_epithelial.rds")


```

### Visualize cluster similarity (PCA embedding and variable feature correlation heatmaps)

```{r}
# Average PCA embedding to show relationship to cluster, annotated by lineage
cluster_embedding <- tibble(cluster_l2 = so_epi@meta.data$cluster_l2) %>%
  cbind(so_epi@reductions$pca@cell.embeddings) %>%
  group_by(cluster_l2) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(data.frame(cluster_embedding[,2:(ndims + 1)],
                              row.names = cluster_embedding$cluster_l2),
                   main = 'Average pca embedding per cluster \n column z-scored',
                   breaks = seq(from = -5, to = 5, length.out = 100))

# Correlation heatmap

Idents(so_epi) <- 'cluster_l2'

av.exp <- AverageExpression(so_epi)$RNA[VariableFeatures(so_epi),]
cor.exp <- as.data.frame(cor(as.matrix(av.exp), method = 'pearson'))

pheatmap::pheatmap(cor.exp)
```


## Differential expression and Enrichment of Gene Ontology

### Find DEGs from subclusters
```{r}
so_epi <- JoinLayers(so_epi)

```


```{r}
if(file.exists('analysis_files/s4_epithelial_markers.csv')){
  print('Epithelial marker csv loading')
  epi_markers <- read_csv('analysis_files/s4_epithelial_markers.csv')
}else{
  print('Computing Epithelial markers then saving')
  Idents(so_epi) <- 'cluster_l2'
  epi_markers <- FindAllMarkers(so_epi,
                                only.pos = TRUE)
  
  write_csv(epi_markers, 'analysis_files/s4_epithelial_markers.csv')
}


```



### Visualize top DEGs

```{r, fig.asp = 1}
epi_top <- epi_markers %>%
  group_by(cluster) %>%
  filter(p_val_adj <= 0.01) %>%
  filter(pct.1 > 0.25) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 5)

so_epi <- ScaleData(so_epi,
                    features = unique(epi_top$gene))

# Heatmap of top markers
DoHeatmap(so_epi,
          features = unique(epi_top$gene))

DotPlot(so_epi,
        features = unique(epi_top$gene),
        cols = 'RdYlBu')+
  coord_flip()+
  RotatedAxis()
```

## EGO

```{r}
subset_ego(curr_sub = so_epi,
           curr_markers = epi_markers)
```


## Define epithelial subtype annotations based on EGO + marker analysis
```{r}
# Define epithelial subtype annotations based on EGO + marker analysis
epi_annotations <- c(
  "epi_0" = "angiogenesis_regulation",
  "epi_1" = "oxphos ",
  "epi_2" = "epiderms_development",
  "epi_3" = "gland_development",
  "epi_4" = "Tcell_activation",
  "epi_5" = "myeloid_chemotaxis",
  "epi_6" = "proliferative",
  "epi_7" = "myeloid_chemotaxis",
  "epi_8"= "proliferative",
  "epi_9"= "ECM_modulation",
  "epi_10" = "fatty_acid_metabolism"
  # adjust cluster IDs and names based on your EGO results
)

# Assign annotations to Seurat object
so_epi$epi_subtype <- dplyr::recode(
  as.character(so_epi$seurat_clusters),
  !!!epi_annotations
)

# Set as identity
Idents(so_epi) <- "epi_subtype"

# Verify
table(so_epi$epi_subtype)

# Visualize
DimPlot(so_epi,
        group.by = "epi_subtype",
        label = FALSE,
        repel = TRUE) +
  ggtitle("Epithelial Subtypes")
```

```{r}
# ── Global phenotype palette (run once at the top of your script) ─────────────
phenotype_colors <- c(
  "PD" = "#7CAE00",   # Combo
  "R"  = "#F8766D",   # PARPI_R
  "S"  = "#00A9FF",   # PARPI_S
  "DT" = "#C77CFF",   # PP2A
  "V"  = "#666666"    # Vehicle
)

phenotype_labels <- c(
  "PD" = "Combo",
  "R"  = "PARPI_R",
  "S"  = "PARPI_S",
  "DT" = "PP2A",
  "V"  = "Vehicle"
)
```


```{r}
so_meta_epi <- so_epi@meta.data

so_meta_tumors_epi <- so_meta_epi %>%
  dplyr::select(epi_subtype, phenotype, tumor) %>%
  group_by(tumor, epi_subtype, phenotype) %>%
  summarize(n_cells = n(), .groups = "drop") %>%
  group_by(tumor) %>%
  mutate(freq = n_cells / sum(n_cells))

so_meta_tumors_summary_epi <- so_meta_tumors_epi %>%
  group_by(phenotype, epi_subtype) %>%
  summarize(
    tumor_count   = n(),
    mean_freq     = mean(freq),
    freq_sd_ymin  = mean(freq) - sd(freq),
    freq_sd_ymax  = mean(freq) + sd(freq),
    freq_sem_ymin = mean(freq) - sd(freq) / (tumor_count)^0.5,
    freq_sem_ymax = mean(freq) + sd(freq) / (tumor_count)^0.5,
    .groups = "drop"
  )

ggplot(so_meta_tumors_summary_epi,
       aes(x = epi_subtype, y = mean_freq, fill = phenotype)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = freq_sem_ymin, ymax = freq_sem_ymax),
                position = "dodge", width = 0.5) +
  facet_grid(rows = vars(phenotype),
             scales = "free_x", space = "free_x") +
  scale_fill_manual(values = phenotype_colors, labels = phenotype_labels) +
  theme_bw() +
  theme(strip.text.y.right = element_text(angle = 0)) +
  xlab("Epithelial Subtype") +
  ylab("Mean frequency") +
  RotatedAxis()
```


# Lymphoid (04/17/26)

## Process lymphoid data


##faster
```{r}
# so_lym (04/17/26)

if (file.exists("analysis_files/s4_so_lymphoid.rds")) {

  message("lymphoid subset already processed — loading")
  so_lym <- readRDS("analysis_files/s4_so_lymphoid.rds")

} else {

  message("Processing lymphoid subset (optimized)")

  DefaultAssay(so_lym) <- "RNA"

  # 1. Normalize (fast)
  so_lym <- NormalizeData(so_lym, verbose = FALSE)

  # 2. Variable features (LIMITED)
  so_lym <- FindVariableFeatures(
    so_lym,
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
  )

  # 3. Scale ONLY variable genes (huge speedup)
  so_lym <- ScaleData(
    so_lym,
    features = VariableFeatures(so_lym),
    verbose = FALSE
  )

  # 4. PCA (few dims first)
  so_lym <- RunPCA(
    so_lym,
    features = VariableFeatures(so_lym),
    npcs = 20,
    verbose = FALSE
  )

  # 5. Neighbors + clustering (lightweight)
  so_lym <- FindNeighbors(so_lym, dims = 1:10, verbose = FALSE)
  so_lym <- FindClusters(so_lym, resolution = 0.4, verbose = FALSE)

  # 6. UMAP (light)
  so_lym <- RunUMAP(so_lym, dims = 1:10, verbose = FALSE)

  saveRDS(so_lym, "analysis_files/s4_so_lymphoid.rds")
}

```

- Load from saved RDS if it exists
- Else: Process subset with 10 PCs

```{r, fig.asp = 1}

if(file.exists('analysis_files/s4_so_lymphoid.rds')){
  print('lymphoid subset already processed - loading saved file')
  so_lym <- readRDS('analysis_files/s4_so_lymphoid.rds')
}else{
  print('Preprocessed lymphoid seurat object not found, processing now. ETA <5 minutes')
  so_lym <- process_subset(so_lym)
}

ElbowPlot(so_lym)

DimPlot(so_lym,
        group.by = 'phenotype')+
  coord_equal()+
  facet_wrap(~phenotype)

DimPlot(so_lym,
        group.by = 'cluster_l1')+
  coord_equal()

# DimPlot(so_lym,
#         group.by = 'Phase')+
#   coord_equal()

```

## Optimize clusters

### Cluster sweep
```{r}
if(file.exists('analysis_files/s4_so_lymphoid.rds')){
  print('lymphoid subset seurat file already exists, not running cluster optimization')
}else{

  so_lym <- FindClusters(
    object = so_lym,
    resolution = seq(from = 0.05, to = 1, by = 0.05),
    algorithm = 3
  )
  
  ndims <- 10
  
  so_cluster_qc <- cluster_qc(seurat_object = so_lym,
                              embedding = so_lym@reductions$pca@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  best_res <- so_cluster_qc %>%
    filter(mean_sil_width/mean_rmsd == max(mean_sil_width/mean_rmsd)) %>%
    dplyr::select(res) %>%
    pull()
  
  so_cluster_qc <- cluster_qc(seurat_object = so_lym,
                              embedding = so_lym@reductions$pca@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  p1 <- ggplot(cluster_qc_gathered, aes(x = res, y = value, color = metric))+
    geom_point()+
    geom_line()+
    theme_bw()+
    facet_wrap(~metric, ncol = 3, scales = 'free_y')+
    scale_x_continuous(breaks = seq(from = 0, to = 2.5, by = 0.5))+
    xlab('Louvain resolution')+
    geom_vline(xintercept = best_res, linetype = 'dashed')
  
  p2 <- ggplot(so_cluster_qc, aes(x = mean_sil_width, y = mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  p3 <- ggplot(so_cluster_qc, aes(x = n_clusters, y = mean_sil_width/mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  plot(p1)
  plot(p2)
  plot(p3)
}

```

### Remove swept clusters and set to optimal resolution

```{r, fig.asp = 1}
# best_res <- 0.1
# 
# if(file.exists('analysis_files/s4_so_lymphoid.rds')){
#   print('lymphoid subset seurat file already exists, optimizing clustering resolution')
# }else{
#  # Store optimal clustering
#   so_lym@meta.data$cluster_l2 <- paste0('lym_', so_lym@meta.data[,paste0('RNA_snn_res.', best_res)])
#   
#   so_lym@meta.data$seurat_clusters <- so_lym@meta.data$cluster_l2
#   
#   DimPlot(so_lym,
#           group.by = 'cluster_l2',
#           label = TRUE)+
#     coord_equal()+
#     theme(legend.position = 'none')
#   
#   # Remove the remainder
#   so_lym@meta.data <- so_lym@meta.data %>%
#     dplyr::select(-grep(colnames(so_lym@meta.data),
#                         pattern = '^SCT_snn_res*',
#                         value = TRUE))
#   
#   # Save processed object
#   saveRDS(so_lym, file = 'analysis_files/s4_so_lymphoid.rds')
# }
# 
# # Visualize optimal sub
# 
# DimPlot(so_lym,
#         group.by = 'cluster_l2',
#         label = TRUE)+
#   coord_equal()
# 
# DimPlot(so_lym,
#         group.by = 'phenotype',
#         label = TRUE)+
#   coord_equal()
# 
# # Overlap
# 
# cluster_overlap <- so_lym@meta.data %>%
#   dplyr::select(cluster_l1, cluster_l2) %>%
#   table()
# 
# lym_nmi <- aricode::NMI(c1 = so_lym@meta.data$cluster_l1,
#              c2 = so_lym@meta.data$cluster_l2)
# 
# pheatmap::pheatmap(cluster_overlap,
#                    display_numbers = round(cluster_overlap),
#                    main = paste0('lymphoid sub/clustering overlap \n NMI = ', round(lym_nmi,2)))
```

```{r}
best_res <- 0.1

# Store optimal clustering (ALWAYS)
so_lym$cluster_l2 <- paste0(
  "lym_",
 so_lym@meta.data[[paste0("RNA_snn_res.", best_res)]]
)

so_lym$seurat_clusters <- so_lym$cluster_l2
Idents(so_lym) <- "cluster_l2"


DimPlot(
  so_lym,
  group.by = "cluster_l2",
  label = TRUE
) +
  coord_equal() +
  theme(legend.position = "none")


# Overlap

cluster_overlap <- so_lym@meta.data %>%
  dplyr::select(cluster_l1, cluster_l2) %>%
  table()

epi_nmi <- aricode::NMI(c1 = so_lym@meta.data$cluster_l1,
                        c2 = so_lym@meta.data$cluster_l2)

saveRDS(so_lym, "analysis_files/s4_so_lymphoid.rds")
```



### Visualize cluster similarity (PCA embedding and correlation heatmaps)

```{r}
# Average pca embedding to show relationship to cluster, annotated by lineage
cluster_embedding <- tibble(cluster_l2 = so_lym@meta.data$cluster_l2) %>%
  cbind(so_lym@reductions$pca@cell.embeddings) %>%
  group_by(cluster_l2) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(data.frame(cluster_embedding[,2:(ndims+1)],
                              row.names = cluster_embedding$cluster_l2),
                   main = 'Average pca embedding per cluster \n column z-scored')

# Correlation heatmap

Idents(so_lym) <- 'cluster_l2'

av.exp <- AverageExpression(so_lym)$RNA[VariableFeatures(so_lym),]
cor.exp <- as.data.frame(cor(as.matrix(av.exp), method = 'pearson'))

pheatmap::pheatmap(cor.exp)
```


## Differential expression and Enrichment of Gene Ontology

### Find DEGs from subclusters

```{r}
if(file.exists('analysis_files/s4_lymphoid_markers.csv')){
  print('lymphoid marker csv loading')
  lym_markers <- read_csv('analysis_files/s4_lymphoid_markers.csv')
}else{
  print('Computing lymphoid markers then saving')
  Idents(so_lym) <- 'cluster_l2'
  lym_markers <- FindAllMarkers(so_lym,
                                only.pos = TRUE)
  
  write_csv(lym_markers, 'analysis_files/s4_lymphoid_markers.csv')
}


```

### Visualize top DEGs

```{r, fig.asp = 1}
lym_top <- lym_markers %>%
  group_by(cluster) %>%
  filter(pct.1 > 0.2) %>%
  filter(p_val_adj <= 0.01) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 5)

so_lym <- ScaleData(so_lym,
                    features = unique(lym_top$gene))

# Heatmap of top markers
DoHeatmap(so_lym,
          features = unique(lym_top$gene))

DotPlot(so_lym,
        features = unique(lym_top$gene),
        cols = 'RdYlBu')+
  coord_flip()+
  RotatedAxis()
```

## EGO

```{r}
subset_ego(curr_sub = so_lym,
           curr_markers = lym_markers)
```

# Myeloid 

## Process myeloid data
##faster
```{r}
# so_mye (04/17/26)

if (file.exists("analysis_files/s4_so_myeloid.rds")) {

  message("myeloid subset already processed — loading")
  so_mye <- readRDS("analysis_files/s4_so_myeloid.rds")

} else {

  message("Processing myeloid subset (optimized)")

  DefaultAssay(so_mye) <- "RNA"

  # 1. Normalize (fast)
  so_mye <- NormalizeData(so_mye, verbose = FALSE)

  # 2. Variable features (LIMITED)
  so_mye <- FindVariableFeatures(
    so_mye,
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
  )

  # 3. Scale ONLY variable genes (huge speedup)
  so_mye <- ScaleData(
    so_mye,
    features = VariableFeatures(so_mye),
    verbose = FALSE
  )

  # 4. PCA (few dims first)
  so_mye <- RunPCA(
    so_mye,
    features = VariableFeatures(so_mye),
    npcs = 20,
    verbose = FALSE
  )

  # 5. Neighbors + clustering (lightweight)
  so_mye <- FindNeighbors(so_mye, dims = 1:10, verbose = FALSE)
 so_mye <- FindClusters(so_mye, resolution = 0.4, verbose = FALSE)

  # 6. UMAP (light)
 so_mye <- RunUMAP(so_mye, dims = 1:10, verbose = FALSE)

  saveRDS(so_mye, "analysis_files/s4_so_myeloid.rds")
}

```

- Load from saved RDS if it exists
- Else: Process subset with 10 PCs

```{r}

if(file.exists('analysis_files/s4_so_myeloid.rds')){
  print('myeloid subset already processed - loading saved file')
  so_mye <- readRDS('analysis_files/s4_so_myeloid.rds')
}else{
  print('Preprocessed myeloid seurat object not found, processing now. ETA <5 minutes')
  so_mye <- process_subset(so_mye)
}

ElbowPlot(so_mye)

DimPlot(so_mye,
        group.by = 'phenotype')+
  coord_equal()+
  facet_wrap(~phenotype)

DimPlot(so_mye,
        group.by = 'cluster_l1')+
  coord_equal()

# DimPlot(so_mye,
#         group.by = 'Phase')+
#   coord_equal()
```

## Optimize clusters

### Cluster sweep
```{r}
if(file.exists('analysis_files/s4_so_myeloid.rds')){
  print('myeloid subset seurat file already exists, not running cluster optimization')
}else{

  so_mye <- FindClusters(
    object = so_mye,
    resolution = seq(from = 0.05, to = 0.7, by = 0.05),
    algorithm = 3
  )
  
  ndims <- 10
  
  so_cluster_qc <- cluster_qc(seurat_object = so_mye,
                              embedding = so_mye@reductions$pca@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  best_res <- so_cluster_qc %>%
    filter(mean_sil_width/mean_rmsd == max(mean_sil_width/mean_rmsd)) %>%
    dplyr::select(res) %>%
    pull()
  
  so_cluster_qc <- cluster_qc(seurat_object = so_mye,
                              embedding = so_mye@reductions$pca@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  p1 <- ggplot(cluster_qc_gathered, aes(x = res, y = value, color = metric))+
    geom_point()+
    geom_line()+
    theme_bw()+
    facet_wrap(~metric, ncol = 3, scales = 'free_y')+
    scale_x_continuous(breaks = seq(from = 0, to = 2.5, by = 0.5))+
    xlab('Louvain resolution')+
    geom_vline(xintercept = best_res, linetype = 'dashed')
  
  p2 <- ggplot(so_cluster_qc, aes(x = mean_sil_width, y = mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  p3 <- ggplot(so_cluster_qc, aes(x = n_clusters, y = mean_sil_width/mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  plot(p1)
  plot(p2)
  plot(p3)
}

# Override best_res due to prior biological knowledge

best_res <- 0.1
```

### Remove swept clusters and set to optimal resolution

```{r, fig.asp = 1}
# if(file.exists('analysis_files/s4_so_myeloid.rds')){
#   print('myeloid subset seurat file already exists, optimizing clustering resolution')
# }else{
#  # Store optimal clustering
#   so_mye@meta.data$cluster_l2 <- paste0('mye_', so_mye@meta.data[,paste0('SCT_snn_res.', best_res)])
#   
#   so_mye@meta.data$seurat_clusters <- so_mye@meta.data$cluster_l2
#   
#   DimPlot(so_mye,
#           group.by = 'cluster_l2',
#           label = TRUE)+
#     coord_equal()+
#     theme(legend.position = 'none')
#   
#   # Remove the remainder
#   so_mye@meta.data <- so_mye@meta.data %>%
#     dplyr::select(-grep(colnames(so_mye@meta.data),
#                         pattern = '*snn_res*',
#                         value = TRUE))
#   
#   # Save processed object
#   saveRDS(so_mye, file = 'analysis_files/s4_so_myeloid.rds')
# }
# 
# # Visualize optimal sub
# 
# DimPlot(so_mye,
#         group.by = 'cluster_l2',
#         label = TRUE)+
#   coord_equal()
# 
# DimPlot(so_mye,
#         group.by = 'phenotype',
#         label = TRUE)+
#   coord_equal()
# 
# # Overlap
# 
# cluster_overlap <- so_mye@meta.data %>%
#   dplyr::select(cluster_l1, cluster_l2) %>%
#   table()
# 
# mye_nmi <- aricode::NMI(c1 = so_mye@meta.data$cluster_l1,
#              c2 = so_mye@meta.data$cluster_l2)
# 
# pheatmap::pheatmap(cluster_overlap,
#                    display_numbers = round(cluster_overlap),
#                    main = paste0('myeloid sub/clustering overlap \n NMI = ', round(mye_nmi,2)))
```

```{r}
best_res <- 0.1

# Store optimal clustering (ALWAYS)
so_mye$cluster_l2 <- paste0(
  "mye_",
 so_mye@meta.data[[paste0("RNA_snn_res.", best_res)]]
)

so_mye$seurat_clusters <- so_mye$cluster_l2
Idents(so_mye) <- "cluster_l2"


DimPlot(
  so_mye,
  group.by = "cluster_l2",
  label = TRUE
) +
  coord_equal() +
  theme(legend.position = "none")


# Overlap

cluster_overlap <- so_mye@meta.data %>%
  dplyr::select(cluster_l1, cluster_l2) %>%
  table()

mye_nmi <- aricode::NMI(c1 = so_mye@meta.data$cluster_l1,
                        c2 = so_mye@meta.data$cluster_l2)

saveRDS(so_mye, "analysis_files/s4_so_myeloid.rds")
```

### Visualize cluster similarity (PCA embedding and correlation heatmaps)

```{r}
# Average pca embedding to show relationship to cluster, annotated by lineage
cluster_embedding <- tibble(cluster_l2 = so_mye@meta.data$cluster_l2) %>%
  cbind(so_mye@reductions$pca@cell.embeddings) %>%
  group_by(cluster_l2) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(data.frame(cluster_embedding[,2:(ndims+1)],
                              row.names = cluster_embedding$cluster_l2),
                   main = 'Average pca embedding per cluster \n column z-scored')

# Correlation heatmap

Idents(so_mye) <- 'cluster_l2'

av.exp <- AverageExpression(so_mye)$RNA[VariableFeatures(so_mye),]
cor.exp <- as.data.frame(cor(as.matrix(av.exp), method = 'pearson'))

pheatmap::pheatmap(cor.exp)
```

## Differential expression and Enrichment of Gene Ontology

### Find DEGs from subclusters

```{r}
if(file.exists('analysis_files/s4_myeloid_markers.csv')){
  print('Myeloid marker csv loading')
  mye_markers <- read_csv('analysis_files/s4_myeloid_markers.csv')
}else{
  print('Computing Myeloid markers then saving')
  Idents(so_mye) <- 'cluster_l2'
  mye_markers <- FindAllMarkers(so_mye,
                                only.pos = TRUE)
  
  write_csv(mye_markers, 'analysis_files/s4_myeloid_markers.csv')
}


```

### Visualize top DEGs

```{r, fig.asp = 1}
mye_top <- mye_markers %>%
  group_by(cluster) %>%
  filter(pct.1 > 0.2) %>%
  filter(p_val_adj <= 0.01) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 3)

so_mye <- ScaleData(so_mye,
                    features = unique(mye_top$gene))

# Heatmap of top markers
DoHeatmap(so_mye,
          features = unique(mye_top$gene))

DotPlot(so_mye,
        features = unique(mye_top$gene),
        cols = 'RdYlBu')+
  coord_flip()+
  RotatedAxis()
```


## EGO

```{r}
subset_ego(curr_sub = so_mye,
           curr_markers = mye_markers)
```

# Stroma (04/17/26)

## Process stroma data

- Load from saved RDS if it exists
- Else: Process subset with 10 PCs
##faster
```{r}
# so_str (04/17/26)

if (file.exists("analysis_files/s4_so_stromal.rds")) {

  message("stromal subset already processed — loading")
  so_str <- readRDS("analysis_files/s4_so_stromal.rds")

} else {

  message("Processing stromal subset (optimized)")

  DefaultAssay(so_str) <- "RNA"

  # 1. Normalize (fast)
  so_str <- NormalizeData(so_str, verbose = FALSE)

  # 2. Variable features (LIMITED)
  so_str <- FindVariableFeatures(
    so_str,
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
  )

  # 3. Scale ONLY variable genes (huge speedup)
  so_str <- ScaleData(
    so_str,
    features = VariableFeatures(so_str),
    verbose = FALSE
  )

  # 4. PCA (few dims first)
  so_str <- RunPCA(
    so_str,
    features = VariableFeatures(so_str),
    npcs = 20,
    verbose = FALSE
  )

  # 5. Neighbors + clustering (lightweight)
  so_str <- FindNeighbors(so_str, dims = 1:10, verbose = FALSE)
 so_str <- FindClusters(so_str, resolution = 0.4, verbose = FALSE)

  # 6. UMAP (light)
 so_str <- RunUMAP(so_str, dims = 1:10, verbose = FALSE)

  saveRDS(so_str, "analysis_files/s4_so_stromal.rds")
}

```

```{r}

if(file.exists('analysis_files/s4_so_stromal.rds')){
  print('stromal subset already processed - loading saved file')
  so_str <- readRDS('analysis_files/s4_so_stromal.rds')
}else{
  print('Preprocessed stromal seurat object not found, processing now. ETA <5 minutes')
  so_str <- process_subset(so_str)
}

ElbowPlot(so_str)

DimPlot(so_str,
        group.by = 'phenotype')+
  coord_equal()+
  facet_wrap(~phenotype)

DimPlot(so_str,
        group.by = 'cluster_l1')+
  coord_equal()

# DimPlot(so_str,
#         group.by = 'Phase')+
#   coord_equal()
```

## Optimize clusters

### Cluster sweep
```{r}
if(file.exists('analysis_files/s4_so_stromal.rds')){
  print('stromal subset seurat file already exists, not running cluster optimization')
}else{

  so_str <- FindClusters(
    object = so_str,
    resolution = seq(from = 0.05, to = 0.7, by = 0.05),
    algorithm = 3
  )
  
  ndims <- 10
  
  so_cluster_qc <- cluster_qc(seurat_object = so_str,
                              embedding = so_str@reductions$pca@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  best_res <- so_cluster_qc %>%
    filter(mean_sil_width/mean_rmsd == max(mean_sil_width/mean_rmsd)) %>%
    dplyr::select(res) %>%
    pull()
  
  so_cluster_qc <- cluster_qc(seurat_object = so_str,
                              embedding = so_str@reductions$pca@cell.embeddings[,1:ndims])
  
  cluster_qc_gathered <- so_cluster_qc %>%
    gather(- c(res), value = 'value', key = 'metric')
  
  p1 <- ggplot(cluster_qc_gathered, aes(x = res, y = value, color = metric))+
    geom_point()+
    geom_line()+
    theme_bw()+
    facet_wrap(~metric, ncol = 3, scales = 'free_y')+
    scale_x_continuous(breaks = seq(from = 0, to = 2.5, by = 0.5))+
    xlab('Louvain resolution')+
    geom_vline(xintercept = best_res, linetype = 'dashed')
  
  p2 <- ggplot(so_cluster_qc, aes(x = mean_sil_width, y = mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  p3 <- ggplot(so_cluster_qc, aes(x = n_clusters, y = mean_sil_width/mean_rmsd, label = res, color = n_clusters))+
    geom_point()+
    ggrepel::geom_text_repel()+
    theme_bw()
  
  plot(p1)
  plot(p2)
  plot(p3)
}

```

### Remove swept clusters and set to optimal resolution

```{r, fig.asp = 1}
# if(file.exists('analysis_files/s4_so_stromal.rds')){
#   print('stromal subset seurat file already exists, optimizing clustering resolution')
# }else{
#  # Store optimal clustering
#   so_str@meta.data$cluster_l2 <- paste0('str_', so_str@meta.data[,paste0('SCT_snn_res.', best_res)])
#   
#   so_str@meta.data$seurat_clusters <- so_str@meta.data$cluster_l2
#   
#   DimPlot(so_str,
#           group.by = 'cluster_l2',
#           label = TRUE)+
#     coord_equal()+
#     theme(legend.position = 'none')
#   
#   # Remove the remainder
#   so_str@meta.data <- so_str@meta.data %>%
#     dplyr::select(-grep(colnames(so_str@meta.data),
#                         pattern = '*snn_res*',
#                         value = TRUE))
#   
#   # Save processed object
#   saveRDS(so_str, file = 'analysis_files/s4_so_stromal.rds')
# }
# 
# # Visualize optimal sub
# 
# DimPlot(so_str,
#         group.by = 'cluster_l2',
#         label = TRUE)+
#   coord_equal()
# 
# DimPlot(so_str,
#         group.by = 'phenotype',
#         label = TRUE)+
#   coord_equal()
# 
# # Overlap
# 
# cluster_overlap <- so_str@meta.data %>%
#   dplyr::select(cluster_l1, cluster_l2) %>%
#   table()
# 
# str_nmi <- aricode::NMI(c1 = so_str@meta.data$cluster_l1,
#              c2 = so_str@meta.data$cluster_l2)
# 
# pheatmap::pheatmap(cluster_overlap,
#                    display_numbers = round(cluster_overlap),
#                    main = paste0('stromal sub/clustering overlap \n NMI = ', round(str_nmi,2)))
```

```{r}
best_res <- 0.1

# Store optimal clustering (ALWAYS)
so_str$cluster_l2 <- paste0(
  "mye_",
 so_str@meta.data[[paste0("RNA_snn_res.", best_res)]]
)

so_str$seurat_clusters <- so_str$cluster_l2
Idents(so_str) <- "cluster_l2"


DimPlot(
  so_str,
  group.by = "cluster_l2",
  label = TRUE
) +
  coord_equal() +
  theme(legend.position = "none")


# Overlap

cluster_overlap <- so_str@meta.data %>%
  dplyr::select(cluster_l1, cluster_l2) %>%
  table()

str_nmi <- aricode::NMI(c1 = so_str@meta.data$cluster_l1,
                        c2 = so_str@meta.data$cluster_l2)

saveRDS(so_str, "analysis_files/s4_so_stromal00.rds")
```


### Visualize cluster similarity (PCA embedding and correlation heatmaps)

```{r}
# Average pca embedding to show relationship to cluster, annotated by lineage
cluster_embedding <- tibble(cluster_l2 = so_str@meta.data$cluster_l2) %>%
  cbind(so_str@reductions$pca@cell.embeddings) %>%
  group_by(cluster_l2) %>%
  summarize(across(everything(), list(mean)))

pheatmap::pheatmap(data.frame(cluster_embedding[,2:(ndims+1)],
                              row.names = cluster_embedding$cluster_l2),
                   main = 'Average pca embedding per cluster \n column z-scored')

# Correlation heatmap

Idents(so_str) <- 'cluster_l2'

av.exp <- AverageExpression(so_str)$RNA[VariableFeatures(so_str),]
cor.exp <- as.data.frame(cor(as.matrix(av.exp), method = 'pearson'))

pheatmap::pheatmap(cor.exp)
```

## Differential expression and Enrichment of Gene Ontology

### Find DEGs from subclusters

```{r}
if(file.exists('analysis_files/s4_stromal_markers.csv')){
  print('stromal marker csv loading')
  str_markers <- read_csv('analysis_files/s4_stromal_markers.csv')
}else{
  print('Computing stromal markers then saving')
  Idents(so_str) <- 'cluster_l2'
  str_markers <- FindAllMarkers(so_str,
                                only.pos = TRUE)
  
  write_csv(str_markers, 'analysis_files/s4_stromal_markers.csv')
}


```

### Visualize top DEGs

```{r, fig.asp = 1}
str_top <- str_markers %>%
  group_by(cluster) %>%
  filter(pct.1 > 0.2) %>%
  filter(p_val_adj <= 0.01) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 5)

so_str <- ScaleData(so_str,
                    features = unique(str_top$gene))

# Heatmap of top markers
DoHeatmap(so_str,
          features = unique(str_top$gene))

DotPlot(so_str,
        features = unique(str_top$gene),
        cols = 'RdYlBu')+
  coord_flip()+
  RotatedAxis()
```

## EGO

```{r}
subset_ego(curr_sub = so_str,
           curr_markers = str_markers)
```


# Canonical marker investigation

## Cytokeratin profile of epithelial clusters

```{r, fig.asp = 1, eval = FALSE}
keratin_features <- grep(row.names(so_epi), pattern = '^Krt', value = TRUE)

keratin_aves <- AverageExpression(so_epi,
                                  features = keratin_features,
                                  assay = 'RNA')

keratin_aves_filtered <- keratin_aves$RNA[rowSums(keratin_aves$RNA) != 0,]
keratin_aves_top <- keratin_aves$RNA[rowSums(keratin_aves$RNA) >= quantile(rowSums(keratin_aves$RNA), 0.9), ]

pheatmap::pheatmap(keratin_aves_filtered)

pheatmap::pheatmap(keratin_aves_top)

DotPlot(so_epi,
        features = row.names(keratin_aves_top),
        cols = 'RdYlBu')+
  coord_flip()+
  RotatedAxis()

rec_goi <- c('Esr1', 'Pgr', 'Erbb2')
lum_goi <- c('Krt7', 'Krt8', 'Krt18', 'Krt19', 'Muc1', 'Mme', 'Cd24a')
bas_goi <- c('Krt5', 'Krt14', 'Krt17', 'Cd44', 'Itga6')
myo_goi <- c('Acta1', 'Acta2', 'Mylk')
immsupp_goi <- c('Pdcd1', 'Pdcd1lg2', 'Ido1', 'Ido2', 'Tgfb1', 'Ccl2', 'Ccl5', 'Gal', 'Cd47', 'Hla-g')
drivers = c('Myc', 'Pten')

DotPlot(so_epi,
        features = list(Subtype = rec_goi,
                        Luminal = lum_goi,
                        Basal = bas_goi,
                        Myoepithelial = myo_goi,
                        Immunosuppressive = immsupp_goi,
                        Drivers = drivers),
        cols = 'RdYlBu',
        scale = FALSE)+
  RotatedAxis()
```

## Lymphoid marker expression

```{r}
lym_goi <- c('Cd3e', 'Cd3g', 'Cd4', 'Cd8a', 'Foxp3', 'Ikzf2', 'Cd19', 'Cd27', 'Igha', 'Sdc1', 'Trdc', 'Mki67', 'Slamf7', 'Cd80', 'Nt5e', 'Pdcd1lg2', 'Ifitm3')

DotPlot(so_lym,
        features = lym_goi,
        cols = 'RdYlBu')+
  RotatedAxis()

DimPlot(so_lym,
        label = TRUE)+
  coord_equal()

# T cell receptor components
tcr_goi <- c('Trac', 'Trdc', 'Trdv4', 'Tcrg-C1')
DotPlot(so_lym,
        features = tcr_goi,
        cols = 'RdYlBu')+
  RotatedAxis()

```

## Myeloid marker expression

```{r}
mye_goi <- c('Ptprc', 'Itgax', 'Itgam', 'Itgae', 'Adgre1', 'Cd14', 'Cd74', 'S100a8', 'S100a9', 'Cd103', 'Cd40', 'Cd80', 'Cd86', 'Ccr7', 'Il10', 'Il12b', 'Tnf', 'Myo1g', 'Cxcl16', 'Adam8', 'Icam1', 'Fscn1', 'Marcks', 'Marcksl1')

DotPlot(so_mye,
        features = mye_goi,
        cols = 'RdYlBu')+
  RotatedAxis()

DimPlot(so_mye,
        label = TRUE)+
  coord_equal()

# mReg vs DC1 vs DC2 markers

dc_goi <- c('Ccr7', 'Fscn1', 'Cd40', 'Cd274', 'Pdcd1lg2', 'Cd200', 'Fas', 'Aldh1a2', 'Il4ra', 'Il4i1', 'Socs2', 'Relb', 'Xcr1', 'Clec9a', 'Cadm1', 'Naaa', 'Cd209a', 'Sirpa', 'H2-DMb2', 'Itgam', 'Itgae', 'Irf7', 'Irf8')

DotPlot(so_mye,
        features = dc_goi,
        cols = 'RdYlBu')+
  RotatedAxis()

DotPlot(so_mye,
        features = list(neut = c('S100a8', 'S100a9', 'Retnlg', 'Mpo', 'Elane'),
                        mono = c('Lyz2', 'Fcgr1', 'Ccr2', 'Cx3cr1', 'S100a4', 'S100a6'),
                        Macro = c('Adgre1', 'Cd68', 'Lgals3', 'Mertk', 'C1qa', 'C1qb'),
                        DC = c('Xcr1', 'Clec9a', 'Batf3', 'Irf8', 'Cdc42', 'Sirpa', 'Itgam', 'Irf4', 'Siglech', 'Bst2')),
        cols = 'RdYlBu',
        cluster.idents = TRUE)+
  RotatedAxis()

```

## Stromal canonical markers

```{r}
str_goi <- c('Acta1', 'Acta2', 'Fap', 'Pdgfra', 'Pdgfrb', 'Pecam1', 'Il6', 'Il11', 'Il17b', 'Dpp4', 'Spp1')

DotPlot(so_str,
        features = str_goi,
        cols = 'RdYlBu')+
  RotatedAxis()
```


# Generate or read cluster_l2 dictionary

```{r}
if(!file.exists('s4_cl2_dict.csv')){
  print('Creating empty cluster_l2<->celltype dictionary - needs to be filled in with celltype labels')
  cl2_dict <- tibble(cluster_l2 = c(unique(so_epi@meta.data$cluster_l2),
                                    unique(so_lym@meta.data$cluster_l2),
                                    unique(so_mye@meta.data$cluster_l2),
                                    unique(so_str@meta.data$cluster_l2)),
                   celltype = '') %>%
    arrange(cluster_l2)
  
  write_csv(cl2_dict, 's4_cl2_dict.csv')
}else{
  print('Loading cluster_l2<->celltype dictionary')
  cl2_dict <- read_csv('s4_cl2_dict.csv')
}
```


## Use celltype dictionary to label each cluster_l2 then save .rds

```{r}
# Assemble barcode <-> cluster_l2 assignments
epi_ids <- tibble(cell_barcode = row.names(so_epi@meta.data),
                  cluster_l2 = so_epi@meta.data$cluster_l2)

lym_ids <- tibble(cell_barcode = row.names(so_lym@meta.data),
                  cluster_l2 = so_lym@meta.data$cluster_l2)

mye_ids <- tibble(cell_barcode = row.names(so_mye@meta.data),
                  cluster_l2 = so_mye@meta.data$cluster_l2)

str_ids <- tibble(cell_barcode = row.names(so_str@meta.data),
                  cluster_l2 = so_str@meta.data$cluster_l2)

# Combine
full_ids <- rbind(epi_ids,
                  lym_ids,
                  mye_ids,
                  str_ids)

# Add celltype label
full_ids$celltype_l2 <- plyr::mapvalues(x = full_ids$cluster_l2,
                                     from = cl2_dict$cluster_l2,
                                     to = cl2_dict$celltype_l2)

full_ids$celltype_l1 <- plyr::mapvalues(x = full_ids$cluster_l2,
                                     from = cl2_dict$cluster_l2,
                                     to = cl2_dict$celltype_l1)

# Join to so_merge@meta.data

new_meta <- left_join(x = so_merge@meta.data,
                                y = full_ids,
                                by = 'cell_barcode')

row.names(new_meta) <- row.names(so_merge@meta.data)

so_merge@meta.data <- new_meta

Idents(so_merge) <- 'celltype_l2'

saveRDS(so_merge, file = 'analysis_files/s4_so_merge_celltypes.rds')
```


# sessionInfo()

```{r}
sessionInfo()
```

