---
title: "s5_17_4_26_ Combination vs solo treatment"
author: "zod"
date: "17/04/2026"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```

<style>
    body .main-container {
        max-width: 2160px;
    }
</style>

# Experiment & contact info

PIs: Rosalie Sears (searsr@ohsu.edu), Laura Heiser (heiserl@ohsu.edu)

Sample preparation: Zinab Doha (dohaz@ohsu.edu)

Library prep from single cells: Xi Li & Colin Daniel (danielc@ohsu.edu)

Analysis: Nick Calistri (calistri@ohsu.edu) and Zinab Doha (dohaz@ohsu.edu)

Sequencing performed by OHSU MPSSR

## Analysis design

- load rliger integrated data (s3_celltypes.rds)

- separate each lineage with multiple clusters (epithelial, fibroblast, lymphoid, myeloid)

- perform DE within each lineage

- Convert cluster to celltype and annotate with relevant pathway/expression 

- Save integrated seurat obejct (s4_so_merge.rds) as well as the lineage subset seurat objects


### Object names

epi = epithelial
sni = stromal non-immune
lym = lymphoid
mye = myeloid

# Set up

## Load libraries
```{r}
library(Matrix)
library(tidyverse)
library(Seurat)
library(clusterProfiler)
library(org.Mm.eg.db)
library(future)
plan("multisession",
     workers = 12)
options(future.globals.maxSize = 25000 * 1024^2) # Sets future maxsize to 2500 MiB

library(rliger)
#library(SeuratWrappers)
library(UpSetR)
library(eulerr)
```

## Set seed

```{r}
set.seed(4)
```


## load s3_celltypes.rds

```{r}
so_merge <- readRDS('analysis_files/s3_lineages_250505.rds')

# Consider batch the vial of tubes that was split across two 10X lanes (prefix of library_id)

so_merge@meta.data$batch <- str_split(so_merge@meta.data$library_id,
                                      pattern = '_',
                                      simplify = TRUE)[,1]

# Remove any previous clustering attempts
so_merge@meta.data <- so_merge@meta.data %>%
  dplyr::select(-grep(colnames(so_merge@meta.data),
                      pattern = '*snn_res*',
                      value = TRUE))

```

## Create subset seurat objects

```{r}
so_epi <- subset(so_merge, subset = lineage == 'epithelial')

so_lym <- subset(so_merge, subset = lineage == 'lymphoid')

so_mye <- subset(so_merge, subset = lineage == 'myeloid')

so_str <- subset(so_merge, subset = lineage == 'stromal')

# Set default ndims
ndims <- 10
```


# Set up metadata colors
```{r}

phenotype_rename <- tibble(shortname = c('V', 'R', 'S', 'DT', 'PD'),
                           longname = c('Vehicle', 'PARPi_R', 'PARPi_S', 'PP2A', 'Combo'))

so_merge@meta.data$phenotype <- plyr::mapvalues(x = so_merge@meta.data$phenotype,
                                              from = phenotype_rename$shortname,
                                               to = phenotype_rename$longname)


col_dic <- tibble(names = c(  'Combo', 'PARPi_R', 'PARPi_S','PP2A' , 'Vehicle'),
                   colors = c( '#7CAE00', '#F8766D', '#00A9FF', '#C77CFF' ,   '#666666' ))
```



## Prepare DEG function

```{r}
subset_ego <- function(curr_sub, curr_markers){

  if(nrow(curr_markers) > 1){
    
    curr_top <- curr_markers %>%
      filter(avg_log2FC > 0.5) %>%
      group_by(cluster) %>%
      arrange(desc(avg_log2FC)) %>%
      filter(p_val_adj <= 0.05) %>%
      slice_head(n = 5)
    
    DoHeatmap(curr_sub, features = curr_top$gene)
    
    for(i in unique(curr_markers$cluster)){
      
      
      clust_markers <- curr_markers %>%
        filter(cluster == i)
      
      volcano_plot <- ggplot(clust_markers, aes(x = avg_log2FC, y = -log10(p_val_adj), label = gene))+
        geom_point()+
        ggrepel::geom_text_repel()+
        ggtitle(paste0('Cluster: ', i))
      
      print(volcano_plot)
      
      print(paste0('Evaluating cluster: ', i))
      curr_top <- curr_markers %>%
        filter(cluster == i) %>%
        filter(avg_log2FC >= 0.5) %>%
        filter(p_val_adj <= 0.05)
      
      curr_bottom <- curr_markers %>%
        filter(cluster == i) %>%
        filter(avg_log2FC <= -0.5) %>%
        filter(p_val_adj <= 0.05)
    
      ego_top <- enrichGO(gene = curr_top$gene,
                OrgDb = org.Mm.eg.db,
                keyType = 'SYMBOL',
                ont = "ALL",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05)
  
      if(if(!is.null(ego_top)){nrow(ego_top) > 0 & nrow(curr_top) >= 3}){
        ego_top_genelist <- curr_top$avg_log2FC
        names(ego_top_genelist) <- curr_top$gene
        ego_top_genelist <- sort(ego_top_genelist, decreasing = TRUE)
        
        p1 <- dotplot(ego_top, title = paste0('Cluster ', i))
        
        p2 <- cnetplot(ego_top,
                       showCategory = 10,
                       foldChange = ego_top_genelist,
                       colorEdge = TRUE)
        
        p3 <- heatplot(ego_top,
                       showCategory = 10,
                       foldChange = ego_top_genelist)
        
        print(p1)
        print(p2)
        print(p3)
    }else(
      print('No upregulated GOs found')
    )
      
    
    }
      
  }else{
    print('FindAllMarkers did not return DEGs (only one cluster in celltype?)')
  }  
}
```



## Combination vs solo treatment

```{r}
# Subset to MYCi (DT), PARPi resistant (R) or combination (PD) treatment
so_combo <- subset(so_merge, subset = phenotype %in% c('DT', 'PD', 'R'))

# Set identity to treatment/response phenotype
Idents(so_combo) <- 'phenotype'

# Sweep through lineages and compute DEGs
for(i in unique(so_combo@meta.data$lineage)){
  # Subset to just current lineage being tested
  so_curr <- subset(so_combo, subset = lineage == i)
  
  # Find DEGs
  pd_v_r_markers <- FindMarkers(so_curr,
                              ident.1 = 'PD',
                              ident.2 = 'R')
  
  pd_v_dt_markers <- FindMarkers(so_curr,
                              ident.1 = 'PD',
                              ident.2 = 'DT')
  
  # Annotate which lineage these belong to and gene name
  pd_v_r_markers$lineage <- i
  pd_v_r_markers$gene <- rownames(pd_v_r_markers)
  pd_v_dt_markers$lineage <- i
  pd_v_dt_markers$gene <- rownames(pd_v_dt_markers)
  
  # Combine together by gene name
  
  combo_markers <- full_join(x = pd_v_r_markers,
                             y = pd_v_dt_markers,
                             by = 'gene',
                             suffix = c('.pd_vs_r', '.pd_vs_dt'))
  
  # Fill in missing values (if logfc is NA, set to 0, if p val is NA set to 1)
  combo_markers$avg_log2FC.pd_vs_r[is.na(combo_markers$avg_log2FC.pd_vs_r)] <- 0
  combo_markers$avg_log2FC.pd_vs_dt[is.na(combo_markers$avg_log2FC.pd_vs_dt)] <- 0
  combo_markers$p_val_adj.pd_vs_r[is.na(combo_markers$p_val_adj.pd_vs_r)] <- 1
  combo_markers$p_val_adj.pd_vs_dt[is.na(combo_markers$p_val_adj.pd_vs_dt)] <- 1
  
  # bi-Volcano plot
  combo_sig <- combo_markers %>%
    filter(abs(avg_log2FC.pd_vs_r) > 0.5) %>%
    filter(abs(avg_log2FC.pd_vs_dt) > 0.5) %>%
    filter(p_val_adj.pd_vs_r < 0.05) %>%
    filter(p_val_adj.pd_vs_dt < 0.05)
  
  bivolp <- ggplot(combo_markers, aes(x = avg_log2FC.pd_vs_r, y = avg_log2FC.pd_vs_dt, label = gene))+
    geom_point()+
    ggrepel::geom_text_repel(data = combo_sig, aes(x = avg_log2FC.pd_vs_r, y = avg_log2FC.pd_vs_dt, label = gene))+
    geom_point(data = combo_sig, aes(x = avg_log2FC.pd_vs_r, y = avg_log2FC.pd_vs_dt, label = gene), color = 'red')+
    ggtitle(i)+
    theme_bw()+
    xlab('PD vs PARPi Resistant \n Average Log2FC')+
    ylab('PD vs MYCi \n Average Log2FC')+
    geom_abline(slope = 1, linetype = 'dashed')+
    coord_equal()
  
  plot(bivolp)
  
  
  #Append DEG markers to combo_lin_markers
  if(i == unique(so_combo@meta.data$lineage)[1]){
    combo_vs_r <- pd_v_r_markers
    combo_vs_dt <- pd_v_dt_markers
  }else{
    combo_vs_r <- rbind(combo_vs_r,
                        pd_v_r_markers)
    combo_vs_dt <- rbind(combo_vs_dt,
                         pd_v_dt_markers)
  }
}

write_csv(combo_vs_r, 'analysis_files/s4_combo_vs_parpiR_markers.csv')
write_csv(combo_vs_dt, 'analysis_files/s4_combo_vs_myci_markers.csv')
```
```{r}
pdr <- read_csv('analysis_files/s4_combo_vs_parpiR_markers.csv')

 # 2/2/23 - added ego
  curr_top <- pdr %>%
    filter(p_val_adj < 0.05 & avg_log2FC > 0.5)
  
  curr_bottom <- pdr %>%
    filter(p_val_adj < 0.05 & avg_log2FC < -0.5)
  
  
  ego_top <- enrichGO(gene = curr_top$gene,
                OrgDb = org.Mm.eg.db,
                keyType = 'SYMBOL',
                ont = "ALL",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05)
  
  ego_bottom <- enrichGO(gene = curr_bottom$gene,
                OrgDb = org.Mm.eg.db,
                keyType = 'SYMBOL',
                ont = "ALL",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05)
  
  
    if(if(!is.null(ego_top)){nrow(ego_top) > 0 & nrow(curr_top) >= 3}){
        ego_top_genelist <- curr_top$avg_log2FC
        names(ego_top_genelist) <- curr_top$gene
        ego_top_genelist <- sort(ego_top_genelist, decreasing = TRUE)
        
        p1 <- dotplot(ego_top, title = paste0('Cluster ', i))
        
        p2 <- cnetplot(ego_top,
                       showCategory = 10,
                       foldChange = ego_top_genelist,
                       colorEdge = TRUE)
        
        p3 <- heatplot(ego_top,
                       showCategory = 10,
                       foldChange = ego_top_genelist)
        
        print(p1)
        print(p2)
        print(p3)
    }else{
      print('No upregulated GOs found')
    }
  
    if(if(!is.null(ego_bottom)){nrow(ego_bottom) > 0 & nrow(curr_bottom) >= 3}){
        ego_bottom_genelist <- curr_bottom$avg_log2FC
        names(ego_bottom_genelist) <- curr_bottom$gene
        ego_bottom_genelist <- sort(ego_bottom_genelist, decreasing = TRUE)
        
        p4 <- dotplot(ego_bottom, title = paste0('Cluster ', i))
        
        p5 <- cnetplot(ego_bottom,
                       showCategory = 10,
                       foldChange = ego_bottom_genelist,
                       colorEdge = TRUE)
        
        p6 <- heatplot(ego_bottom,
                       showCategory = 10,
                       foldChange = ego_bottom_genelist)
        
        print(p4)
        print(p5)
        print(p6)
    }else{
      print('No upregulated GOs found')
    }
  
  
  # Append DEG markers to parp_lin_markers
  if(i == unique(so_DTV@meta.data$lineage)[1]){
    parp_lin_markers <- curr_markers
  }else{
    parp_lin_markers <- rbind(parp_lin_markers,
                              curr_markers)
  }


# Filter to significant genes
parp_lin_sig_up <- parp_lin_markers %>%
  filter(p_val_adj < 0.01 & avg_log2FC > 0.5)

parp_lin_sig_down <- parp_lin_markers %>%
  filter(p_val_adj < 0.01 & avg_log2FC < -0.5)

# Convert to list for upset plot
parp_up_list <- unstack(parp_lin_sig_up, gene~lineage)
parp_down_list <- unstack(parp_lin_sig_down, gene~lineage) 

# Create upset plot
upset(fromList(parp_up_list))
upset(fromList(parp_down_list))

# Create Eulerr plots

plot(euler(fromList(parp_up_list)),
     quantities = TRUE,
     labels = list(font = 4),
     main = 'Significantly upregulated genes')

plot(euler(fromList(parp_down_list)),
     quantities = TRUE,
     labels = list(font = 4),
     main = 'Significantly downregulated genes')

```

##HR scor
```{r}
hr_genes <- list(HR = c("Brca1", "Brca2", "Rad51", "Rad52", "Palb2",
                         "Rad51c", "Rad51d", "Bard1", "Atm", "Atr",
                         "Chek1", "Chek2", "Fancd2", "Fanca", "Fancl"))

# Check how many are found
hr_genes_present <- hr_genes$HR[hr_genes$HR %in% rownames(so_epi)]
cat("Genes found:", length(hr_genes_present), "/", length(hr_genes$HR), "\n")
cat(hr_genes_present, "\n")
```

```{r}
expr_matrix <- LayerData(so_epi, layer = "data")[hr_genes_present, ]
hr_scores   <- colMeans(as.matrix(expr_matrix))

so_epi@meta.data$HR_score <- hr_scores[rownames(so_epi@meta.data)]

# Rebuild hr_meta
hr_meta <- so_epi@meta.data[, c("phenotype", "HR_score")]
hr_meta <- hr_meta[!is.na(hr_meta$HR_score), ]
hr_meta$phenotype <- factor(hr_meta$phenotype, levels = col_dic$names)

table(hr_meta$phenotype)
```
```{r}
# ── Pairwise wilcox ───────────────────────────────────────────────────────────
pwc <- hr_meta %>%
  pairwise_wilcox_test(HR_score ~ phenotype) %>%
  add_xy_position(x = "phenotype")

# ── Plot ──────────────────────────────────────────────────────────────────────
ggplot(hr_meta, aes(x = phenotype, y = HR_score, fill = phenotype)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               fill = "white", color = "black") +
  geom_jitter(aes(color = phenotype),
              size = 0.3, alpha = 0.3,
              position = position_jitter(width = 0.15)) +
  stat_pvalue_manual(
    filter(pwc, p.adj < 0.05),
    label         = "p.adj",
    tip.length    = 0.01,
    step.increase = 0.08,
    size          = 3.5
  ) +
  scale_fill_manual(values  = setNames(col_dic$colors, col_dic$names)) +
  scale_color_manual(values = setNames(col_dic$colors, col_dic$names)) +
  theme_classic() +
  theme(
    legend.position = "right",
    axis.text.x     = element_text(angle = 45, hjust = 1)
  ) +
  labs(title = "HR module score",
       x     = "Identity",
       y     = "HR module score",
       fill  = NULL, color = NULL)
```

##per tuomr
```{r}
# ── Build per-tumor HR score summary ─────────────────────────────────────────
hr_tumor <- so_epi@meta.data[, c("phenotype", "HR_score", "tumor")]
hr_tumor <- hr_tumor[!is.na(hr_tumor$HR_score), ]
hr_tumor$phenotype <- factor(hr_tumor$phenotype, levels = col_dic$names)

# ── Pairwise wilcox per phenotype (using tumor means) ────────────────────────
hr_tumor_mean <- hr_tumor %>%
  group_by(tumor, phenotype) %>%
  summarize(mean_HR = mean(HR_score), .groups = "drop") %>%
  mutate(phenotype = factor(phenotype, levels = col_dic$names))

# Pairwise test on tumor-level means
pwc_tumor <- hr_tumor_mean %>%
  pairwise_wilcox_test(mean_HR ~ phenotype) %>%
  add_xy_position(x = "phenotype")

# ── Plot ──────────────────────────────────────────────────────────────────────
ggboxplot(hr_tumor_mean,
          x      = "phenotype",
          y      = "mean_HR",
          fill   = "phenotype",
          add    = "jitter",
          width  = 0.5) +
  stat_pvalue_manual(
    filter(pwc_tumor, p.adj < 0.05),
    label         = "p.adj",
    tip.length    = 0.01,
    step.increase = 0.1,
    size          = 3.5
  ) +
  scale_fill_manual(values  = setNames(col_dic$colors, col_dic$names)) +
  scale_color_manual(values = setNames(col_dic$colors, col_dic$names)) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title   = "HR module score per tumor",
    x       = "Phenotype",
    y       = "Mean HR score (per tumor)",
    caption = get_pwc_label(pwc_tumor)
  )

```
##HR genesets in MSigDB (mouse)
```{r}
# ── Install if needed ─────────────────────────────────────────────────────────
# BiocManager::install("msigdbr")
library(msigdbr)

# ── Browse available HR-related genesets ──────────────────────────────────────
# Search for HR genesets in MSigDB (mouse)
all_genesets <- msigdbr(species = "Mus musculus")

hr_sets <- all_genesets %>%
  filter(grepl("HOMOLOGOUS_RECOMBINATION|HALLMARK_DNA_REPAIR|
                FANCONI|RAD51|BRCA", 
               gs_name, ignore.case = TRUE)) %>%
  distinct(gs_name)

print(hr_sets)

#write.csv(hr_sets, "hr_sets.csv")

```


```{r}
# ── Pull the best HR genesets ─────────────────────────────────────────────────

# Option 1: Reactome HRR — most mechanistically curated (recommended)
hr_reactome <- all_genesets %>%
  filter(gs_name == "REACTOME_HDR_THROUGH_HOMOLOGOUS_RECOMBINATION_HRR")

# Option 2: GO Biological Process HR — ontology-based
hr_gobp <- all_genesets %>%
  filter(gs_name == "GOBP_HOMOLOGOUS_RECOMBINATION")

# Option 3: KEGG HR — classic pathway
hr_kegg <- all_genesets %>%
  filter(gs_name == "KEGG_HOMOLOGOUS_RECOMBINATION")

# ── Check gene counts & overlap ───────────────────────────────────────────────
for(gs in list(hr_reactome, hr_gobp, hr_kegg)){
  genes_present <- gs$gene_symbol[gs$gene_symbol %in% rownames(so_epi)]
  cat(unique(gs$gs_name), "→ total:", nrow(gs),
      "| found in so_epi:", length(genes_present), "\n")
}

# ── Use Reactome (recommended) ────────────────────────────────────────────────
hr_genes_msig <- list(HR = hr_reactome$gene_symbol)
hr_genes_present <- hr_genes_msig$HR[hr_genes_msig$HR %in% rownames(so_epi)]
cat("Using genes:\n", paste(hr_genes_present, collapse = ", "), "\n")

# ── Recompute HR score ────────────────────────────────────────────────────────
expr_matrix <- LayerData(so_epi, layer = "data")[hr_genes_present, ]
hr_scores   <- colMeans(as.matrix(expr_matrix))
so_epi@meta.data$HR_score <- hr_scores[rownames(so_epi@meta.data)]

# ── Rebuild hr_meta & hr_tumor_mean ──────────────────────────────────────────
hr_tumor <- so_epi@meta.data[, c("phenotype", "HR_score", "tumor")]
hr_tumor$phenotype <- factor(hr_tumor$phenotype, levels = col_dic$names)

hr_tumor_mean <- hr_tumor %>%
  group_by(tumor, phenotype) %>%
  summarize(mean_HR = mean(HR_score), .groups = "drop")

# ── Test + plot ───────────────────────────────────────────────────────────────
pwc_tumor <- hr_tumor_mean %>%
  pairwise_wilcox_test(mean_HR ~ phenotype) %>%
  add_xy_position(x = "phenotype")

ggboxplot(hr_tumor_mean,
          x     = "phenotype",
          y     = "mean_HR",
          fill  = "phenotype",
          add   = "jitter",
          width = 0.5) +
  stat_pvalue_manual(
    filter(pwc_tumor, p.adj < 0.05),
    label         = "p.adj",
    tip.length    = 0.01,
    step.increase = 0.1,
    size          = 3.5
  ) +
  scale_fill_manual(values  = setNames(col_dic$colors, col_dic$names)) +
  scale_color_manual(values = setNames(col_dic$colors, col_dic$names)) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x     = element_text(angle = 45, hjust = 1)) +
  labs(title   = "HR module score (Reactome HRR)",
       x       = "Phenotype",
       y       = "Mean HR score (per tumor)",
       caption = get_pwc_label(pwc_tumor))
```

```{r}
install.packages("ggpubr")
library(ggpubr)
# ── Fix 1: Join layers in so_epi first ───────────────────────────────────────
so_epi <- JoinLayers(so_epi)
# 
# # ── Fix 2: Load rstatix ──────────────────────────────────────────────────────
 library(rstatix)
# ── GOBP Homologous Recombination geneset ─────────────────────────────────────
hr_gobp <- all_genesets %>%
  filter(gs_name == "GOCC_BRCA1_A_COMPLEX")

cat("Total genes in GOBP_HR:", nrow(hr_gobp), "\n")

# ── Check overlap ─────────────────────────────────────────────────────────────
hr_genes_present <- hr_gobp$gene_symbol[hr_gobp$gene_symbol %in% rownames(so_epi)]
cat("Found in so_epi:", length(hr_genes_present), "/", nrow(hr_gobp), "\n")
cat(hr_genes_present, "\n")

# ── Recompute HR score ────────────────────────────────────────────────────────
expr_matrix <- LayerData(so_epi, layer = "data")[hr_genes_present, ]
hr_scores   <- colMeans(as.matrix(expr_matrix))
so_epi@meta.data$HR_score <- hr_scores[rownames(so_epi@meta.data)]

# ── Rebuild per-tumor means ───────────────────────────────────────────────────
hr_tumor_mean <- so_epi@meta.data[, c("phenotype", "HR_score", "tumor")] %>%
  filter(!is.na(HR_score)) %>%
  mutate(phenotype = factor(phenotype, levels = col_dic$names)) %>%
  group_by(tumor, phenotype) %>%
  summarize(mean_HR = mean(HR_score), .groups = "drop")

# ── Pairwise wilcox ───────────────────────────────────────────────────────────
pwc_tumor <- hr_tumor_mean %>%
  pairwise_wilcox_test(mean_HR ~ phenotype) %>%
  add_xy_position(x = "phenotype")

# ── Plot ──────────────────────────────────────────────────────────────────────
ggboxplot(hr_tumor_mean,
          x     = "phenotype",
          y     = "mean_HR",
          fill  = "phenotype",
          add   = "jitter",
          width = 0.5) +
  stat_pvalue_manual(
    filter(pwc_tumor, p.adj < 0.05),
    label         = "p.adj",
    tip.length    = 0.01,
    step.increase = 0.1,
    size          = 3.5
  ) +
  scale_fill_manual(values  = setNames(col_dic$colors, col_dic$names)) +
  scale_color_manual(values = setNames(col_dic$colors, col_dic$names)) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x     = element_text(angle = 45, hjust = 1)) +
  labs(title   = "HR module score (GOBP_HOMOLOGOUS_RECOMBINATION)",
       x       = "Phenotype",
       y       = "Mean HR score (per tumor)",
       caption = get_pwc_label(pwc_tumor))
```


##cluster_frequency_stats
```{r}
library(ggpubr)
library(rstatix)

# Loop to plot all epi subtypes, or pick one
sub_of_interest <- 'epi_4-Tcell_activation'  # change to any epi_subtype name

soi <- so_meta_tumors_epi %>%
  ungroup() %>%
  filter(epi_subtype == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(phenotype, freq) %>%
  pairwise_t_test(freq ~ phenotype) %>%
  add_xy_position(x = 'phenotype')

ggboxplot(soi, 
          x     = "phenotype", 
          y     = "freq", 
          color = "phenotype", 
          add   = "jitter") +
  scale_color_manual(values = setNames(col_dic$colors, col_dic$names)) +
  stat_pvalue_manual(
    filter(soi_pwc, p.adj < 0.25), 
    label         = "p.adj", 
    tip.length    = 0, 
    step.increase = 0.1
  ) +
  labs(caption = get_pwc_label(soi_pwc)) +
  ggtitle(sub_of_interest) +
  ylab('Frequency of subcluster in lineage (per tumor)') +
  xlab('') +
  theme(legend.position = 'none') +
  RotatedAxis()
```

```{r, eval = FALSE}
saveRDS(so_merge, file = 'analysis_files/s4_so_merge.rds')
```



# sessionInfo()

```{r}
sessionInfo()
```

