---
title: "260120_s6_cluster_frequency_stats"
author: "nlc and zod"
date: "2026-04-20"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

## Script goals

- Read in subclustered seurat object of MycPtenfl tumors from all treatments/phenotypes

- Collapse the Fibroblast, Endothelial and Perivascular lineages into a single 'stromal' lineage

- For each tumor: Compute the proportion of tumor which was subcluster, proportion of lineage which was subcluster

- Perform pairwise t.test across phenotype/treatments to identify subclusters which are significantly enriched in a given phenotype

- General purpose code to plot boxplots with significance visualized

- PCA/UMAP embedding of cluster proportion

- Correlation of cluster proportion to highlight potentially correlated clusters

- 

## Load libraries

```{r}
library(tidyverse)
library(Seurat)
library(rstatix)
library(ggpubr)
```

## Load data

```{r}
so <- readRDS('analysis_files/s4_so_merge_subclusters.rds')

# Recode stromal lineages (combine fibroblast/endothelial/perivascular into just 'stromal')
#so@meta.data$lineage <- ifelse(so@meta.data$lineage %in% c('fibroblast', 'endothelial', 'perivascular'),
                          #     'stromal',
                            #   so@meta.data$lineage)

# Map subcluster <-> lineage
sub_lin_dict <- so@meta.data %>%
  dplyr::select(subcluster, lineage) %>%
  distinct()

# Map tumor <-> lib_exp (library/experiment - the day that experiment was performed representing experimental batch but NOT exactly sequencing library since HTO pooled samples were split on duplicate 10X lanes)
tumor_lib_dict <- so@meta.data %>%
  mutate(lib_exp = str_split(str_remove(library_id, pattern = '../'), pattern = '_', simplify = TRUE)[,1]) %>%
  dplyr::select(tumor, lib_exp) %>%
  distinct()
```

# Set up metadata colors
```{r}

treatment_rename <- tibble(shortname = c('V', 'R', 'S', 'DT', 'PD'),
                           longname = c('Vehicle', 'PARPi_R', 'PARPi_S', 'PP2A', 'Combo'))

so@meta.data$treatment <- plyr::mapvalues(x = so@meta.data$phenotype,
                                              from = treatment_rename$shortname,
                                               to = treatment_rename$longname)


col_dic <- tibble(names = c(  'PARPi_S', 'Combo', 'PARPi_R','PP2A' , 'Vehicle'),
                   colors = c( '#7CAE00', '#F8766D', '#00A9FF', '#C77CFF' ,   '#666666' ))
```

# Pre process

## Compute proportions for each tumor

freq = proportion of tumor (per tumor)

freq_lineage = proportion of lineage (per tumor)

```{r}
so_meta_tumors <- so@meta.data %>%
  dplyr::select(subcluster, phenotype, lineage, tumor) %>%
  group_by(tumor, phenotype, subcluster, lineage) %>%
  summarize(n_cells = n()) %>%
  group_by(tumor) %>%
  mutate(freq = n_cells/sum(n_cells)) %>%
  group_by(tumor, lineage) %>%
  mutate(freq_lineage = n_cells/sum(n_cells))  %>%
  mutate(treatment = plyr::mapvalues(x = phenotype,
                                              from = treatment_rename$shortname,
                                               to = treatment_rename$longname))
  
# Expand all possible subcluster x tumor combinations
subcluster_tumor_expanded <- so_meta_tumors %>%
  ungroup() %>%
  tidyr::expand(subcluster, tumor)

# Combine observations with expanded subcluster x tumor combinations, fill in NAs where appropriate
so_meta_complete <- full_join(x = subcluster_tumor_expanded,
                              y = so_meta_tumors,
                              suffix = c('', ''),
                              by = c('subcluster', 'tumor')) %>%
  mutate(phenotype = ifelse(is.na(phenotype),
                            gsub("[^a-zA-Z]", "", str_split(tumor, pattern = '_', simplify = TRUE)[,1]),
                            phenotype)) %>%
  mutate(lineage = ifelse(is.na(lineage),
                          plyr::mapvalues(x = subcluster,
                                          from = sub_lin_dict$subcluster,
                                          to = sub_lin_dict$lineage),
                          lineage)) %>%
  mutate(n_cells = ifelse(is.na(n_cells), 0, n_cells)) %>%
  mutate(freq = ifelse(is.na(freq), 0, freq)) %>%
  mutate(freq_lineage = ifelse(is.na(freq_lineage), 0, freq_lineage)) %>%
  mutate(lib_exp = plyr::mapvalues(x = tumor,
                                   from = tumor_lib_dict$tumor,
                                   to = tumor_lib_dict$lib_exp))  %>%
  mutate(treatment = plyr::mapvalues(x = phenotype,
                                              from = treatment_rename$shortname,
                                               to = treatment_rename$longname)) %>%
  mutate(treatment = factor(treatment, levels = c('PP2A', 'Combo',  'PARPi_R', 'PARPi_S', 'Vehicle')))

```

## Barplot to visualize data overall

error bar represents SEM

```{r, fig.asp = 1}
n_tumors <- so@meta.data %>%
  group_by(treatment) %>%
  dplyr::select(treatment, tumor) %>%
  distinct() %>%
  summarize(n_tumors = n())


so_meta_tumors_summary <- so_meta_tumors %>%
  mutate(tumor_count  = as.numeric(plyr::mapvalues(x = treatment,
                                                     from = n_tumors$treatment,
                                                     to = n_tumors$n_tumors)))%>%
  group_by(treatment, subcluster, lineage, tumor_count) %>%
  summarize(mean_freq = mean(freq),
            sd_freq = sd(freq))

ggplot(so_meta_tumors_summary, aes(y = subcluster, x = mean_freq, fill = treatment))+
  geom_col(position = 'dodge')+
  theme_bw()+
  geom_errorbar(aes(y = subcluster, xmin = mean_freq - sd_freq/(tumor_count^0.5), xmax = mean_freq + sd_freq/(tumor_count^0.5)), position = 'dodge', width = 0.5)+
  facet_grid(rows = vars(lineage), cols = vars(treatment), scales = 'free_y', space = 'free')+
  theme(strip.text.y.right = element_text(angle = 0))+
  xlab('Mean frequency')+
  ylab('Celltype \n (cluster_lineage_celltype)')
```
# Celltype fractions

## Visualize celltype fractions per tumor
```{r}



so_merge_meta <- so_merge@meta.data %>%
  mutate(tumor_pheno = paste0(tumor, '-', treatment))

freq_table <- so_merge_meta %>%
  dplyr::select(lineage, tumor_pheno) %>%
  table()

col_anno <- data.frame(treatment = str_split(colnames(freq_table), pattern = '-', simplify = TRUE)[,2],
                       log10_n_cells = log10(colSums(freq_table)),
                       row.names = colnames(freq_table))

pheatmap::pheatmap(prop.table(freq_table, margin = 2), 
                   display_numbers = freq_table,
                   main = 'celltype fraction \n Number = n_cells per celltype per tumor',
                   annotation_col = col_anno)
```

# Statistical testing

## (phenotype) Pairwise  t test for proportion of tumor

```{r}
pwc_all <- so_meta_complete %>%
  ungroup() %>%
  group_by(subcluster) %>%
  dplyr::select(subcluster, freq, treatment) %>%
  pairwise_t_test(freq~treatment)

write_csv(pwc_all, 'pwt_all.csv')

pwc_all %>%
  filter(p.adj < 0.05) %>%
  knitr::kable()
```

## (phenotype) Pairwise t.test for proportion of lineage

```{r}
pwc_all_lin <- so_meta_complete %>%
  ungroup() %>%
  group_by(subcluster) %>%
  dplyr::select(subcluster, freq_lineage, treatment) %>%
  pairwise_t_test(freq_lineage~treatment)

write_csv(pwc_all_lin, 'pwt_all_lineage.csv')

pwc_all_lin %>%
  filter(p.adj < 0.05) %>%
  knitr::kable()
```


## (lib_exp) Pairwise t test for proportion of tumor

```{r}
pwc_all_lib_exp <- so_meta_complete %>%
  ungroup() %>%
  group_by(subcluster) %>%
  dplyr::select(subcluster, freq, lib_exp) %>%
  pairwise_t_test(freq~lib_exp)

pwc_all_lib_exp %>%
  filter(p.adj < 0.05) %>%
  knitr::kable()
```

## (phenotype) Pairwise t.test for proportion of lineage

```{r}
pwc_all_lin_lib_exp <- so_meta_complete %>%
  ungroup() %>%
  group_by(subcluster) %>%
  dplyr::select(subcluster, freq_lineage, lib_exp) %>%
  pairwise_t_test(freq_lineage~lib_exp)


pwc_all_lin_lib_exp %>%
  filter(p.adj < 0.05) %>%
  knitr::kable()
```

# Boxplot with stats

## Proportion of tumor

```{r}
sub_of_interest <- 'lym_7'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter')+ scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666')) +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```



## Proportion of lineage

```{r}



sub_of_interest <- 'lym_7'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')  
```
#lym_0
## Proportion of tumor

```{r}
sub_of_interest <- 'lym_0'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```



## Proportion of lineage

```{r}
sub_of_interest <- 'lym_0'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
#mye_o
## Proportion of tumor

```{r}
sub_of_interest <- 'mye_0'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```



## Proportion of lineage

```{r}
sub_of_interest <- 'mye_0'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
#mye_3
## Proportion of tumor

```{r}
sub_of_interest <- 'mye_3'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```



## Proportion of lineage

```{r}
sub_of_interest <- 'mye_3'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
#mye_4
## Proportion of tumor

```{r}
sub_of_interest <- 'mye_4'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```



## Proportion of lineage

```{r}
sub_of_interest <- 'mye_4'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```

## epi4
## Proportion of tumor

```{r}
sub_of_interest <- 'epi_4'

soi <- so_meta_complete %>%
  ungroup() %>%
 filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+ 
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'epi_4'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
## epi5
## Proportion of tumor

```{r}
sub_of_interest <- 'epi_5'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +  scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'epi_5'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
## epi1
## Proportion of tumor

```{r}
sub_of_interest <- 'epi_1'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'epi_1'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
## epi9
## Proportion of tumor

```{r}
sub_of_interest <- 'epi_9'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'epi_9'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
## epi10
## Proportion of tumor

```{r}
sub_of_interest <- 'epi_10'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'epi_10'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
## str_5
## Proportion of tumor

```{r}
sub_of_interest <- 'str_5'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq) %>%
  pairwise_t_test(freq~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq", color = 'treatment', add = 'jitter') +
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in tumor')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'str_5'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```
## Proportion of lineage

```{r}
sub_of_interest <- 'str_6'

soi <- so_meta_complete %>%
  ungroup() %>%
  filter(subcluster == sub_of_interest)

soi_pwc <- soi %>%
  dplyr::select(treatment, freq_lineage) %>%
  pairwise_t_test(freq_lineage~treatment) %>%
  add_xy_position(x = 'treatment')

ggboxplot(soi, x = "treatment", y = "freq_lineage", color = 'treatment', add = 'jitter') + scale_color_manual(values=c('#C77CFF', '#7CAE00', '#F8766D', '#00A9FF' ,  '#666666'))+
  stat_pvalue_manual(filter(soi_pwc, p.adj < 0.25), label = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(caption = get_pwc_label(soi_pwc))+
  ggtitle(sub_of_interest)+
  ylab('Frequency of subcluster in lineage (per tumor)')
```


# Additional plots

## Scatter plot of freq vs freq_lineage for each subcluster

```{r, fig.asp = 2}
ggplot(so_meta_tumors, aes(x = freq, y = freq_lineage, color = phenotype))+
  geom_point()+
  facet_wrap(~subcluster, scale = 'free', ncol = 5)+
  theme_bw()+
  RotatedAxis()

so_meta_tumors$lib_exp <- plyr::mapvalues(x = so_meta_tumors$tumor,
                                            from = tumor_lib_dict$tumor,
                                            to = tumor_lib_dict$lib_exp)

ggplot(so_meta_tumors, aes(x = freq, y = freq_lineage, color = lib_exp))+
  geom_point()+
  facet_wrap(~subcluster, scale = 'free', ncol = 5)+
  theme_bw()+
  RotatedAxis()

```

## PCA of proportion of tumor

```{r}
freqs <- so@meta.data %>%
  dplyr::select(tumor, subcluster) %>%
  table()

props <- freqs %>%
  prop.table(., margin = 1)

pc.props <- prcomp(x = props)

pc.tibble <- tibble(tumor = rownames(pc.props$x),
                    PC_1 = pc.props$x[,1],
                    PC_2 = pc.props$x[,2]) %>%
  mutate(phenotype = gsub("[^a-zA-Z]", "", str_split(tumor, pattern = '_', simplify = TRUE)[,1])) %>%
  mutate(library = plyr::mapvalues(x = tumor,
                                   from = tumor_lib_dict$tumor,
                                   to = tumor_lib_dict$lib_exp))

ggplot(pc.tibble, aes(x = PC_1, y = PC_2, color = phenotype))+
  geom_point(size = 3)+
  theme_bw()+
  coord_equal()

ggplot(pc.tibble, aes(x = PC_1, y = PC_2, color = library))+
  geom_point(size = 3)+
  theme_bw()+
  coord_equal()
```

## UMAP of proportion of tumor

```{r}
library(umap)

config.custom <- umap.defaults
config.custom$n_neighbors <- 5

npcs <- which.min(as.data.frame(summary(pc.props)$importance)[3,] < 0.95)

umap.props <- umap(d = pc.props$x[,1:npcs],
                   config = config.custom)

umap.tibble <- as_tibble(umap.props$layout) %>%
  mutate(tumor = row.names(umap.props$layout)) %>%
  mutate(phenotype = gsub("[^a-zA-Z]", "", str_split(tumor, pattern = '_', simplify = TRUE)[,1])) %>%
  mutate(library = plyr::mapvalues(x = tumor,
                                   from = tumor_lib_dict$tumor,
                                   to = tumor_lib_dict$lib_exp))

ggplot(umap.tibble, aes(x = V1, y = V2, color = phenotype))+
  geom_point(size = 3)+
  theme_bw()+
  coord_equal()

ggplot(umap.tibble, aes(x = V1, y = V2, color = library))+
  geom_point(size = 3)+
  theme_bw()+
  coord_equal()
```

# Correlation of all subclusters

```{r, fig.asp = 1}
# Correlation of clusters
pd_v_r_sig <- pwc_all %>%
  filter(group1 == 'PD' & group2 == 'R' & p.adj <= 0.05)

pd_v_v_sig <- pwc_all %>%
  filter(group1 == 'PD' & group2 == 'V' & p.adj <= 0.05)
dt_v_v_sig <- pwc_all %>%
  filter(group1 == 'DT' & group2 == 'V' & p.adj <= 0.05)
r_v_v_sig <- pwc_all %>%
  filter(group1 == 'R' & group2 == 'V' & p.adj <= 0.05)


cor.props <- cor(props)

col_anno <- data.frame(row.names = row.names(cor.props),
                       PD_v_R_sig = 1*(row.names(cor.props) %in% pd_v_r_sig$subcluster))

row_anno <- data.frame(row.names = row.names(cor.props),
                       PD_v_V_sig = 1*(row.names(cor.props) %in% pd_v_v_sig$subcluster),
                       DT_v_V_sig = 1*(row.names(cor.props) %in% dt_v_v_sig$subcluster),
                       R_v_V_sig = 1*(row.names(cor.props) %in% r_v_v_sig$subcluster))

pheatmap::pheatmap(cor.props,
                   cutree_rows = 4,
                   cutree_cols = 4,
                   annotation_col = col_anno,
                   annotation_row = row_anno)

```



# sessionInfo()

```{r}
sessionInfo()
```


# 